# (PART) Frameworks {.unnumbered}

# Introduction {#intro}

## Background of the study

-   Studying social hierarchy in Prehistory through fractal analysis of settlement plans

-   In Europe, the Neolithic is the long and messy transition period between mobile hunter-gatherer groups in the Palaeo- and Mesolithic, and the first city states in the Bronze (Aegean) and Iron Ages (Mediterranean and Central Europe)

-   Early farming economy, influence from the Near East.

-   Large variety in scale and content of archaeologically defined culture groups. Single farmsteads and small hamlets in many phases -- some with hardly any settlement evidence at all **examples mid-neo, use** @shennan2018. Other phases include exceptionally large settlements, probably hosting populations of several thousand inhabitants, like at Maidanetske in central Ukraine around 3.800 BCE (see \@ref(trypillia)). While in some settings, like in Linear Pottery society in much of continental Europe north of the Alps towards 5.100 BCE (see \@ref(lbk)), the dead were buried in simple pit graves, with very little distinction in treatment between individuals, in other phases some individuals were buried with tremendous amounts of precious goods like in Varna, Romania, or under colossal burial mounds like in Carnac, France, both in the mid-5th millennium [@shennan2018]**check ref**.

-   Seen at a very large scale -- across the continent and through the Holocene -- the development of society from small scale and relatively egalitarian towards large scale and more hierarchical seems evident (though not to everyone, see @graeber2021 and Section \@ref(soc-typ)). When we look more closely however, this evolution is anything but linear, as both population sizes and levels of hierarchical organisation seem to fluctuate considerably, sometimes over short time spans as from the Trypillia **C2 to D1??, check this**, when the so-called mega-sites are abandoned and their former inhabitants regroup into much smaller settlements during a transition of maybe only **check and ref** years.

-   In many cases, the level of social hierarchisation and complexity in a given Prehistoric society is very hard for researchers to evaluate, since many indicators of such structures are either lost from the archaeological record, or were never included in the first place [@perreault2019; Section \@ref(complexity-archaeo)]. Archaeological traces that are often interpreted as signs of social complexity and hierarchy may furthermore be deceiving. Seemingly monumental structures were in many cases built through small additions over centuries, rather than in one colossal construction campaign [**example, Carnac alignments? Danish megaliths?**]. In many megalithic burial contexts, it may be impossible to know how large a segment of the society that had access to such inhumations (**rephrase**).

-   New methods for investigating hierarchy: fractal analysis. Borrowed from other disciplines, not much tested in Archaeology.

## Research question and objectives

The overall goal of the present study is to test and assess the utility of fractal analysis techniques as tools for studying hierarchical social organisation in prehistoric societies. Two methodological approaches are under special scrutiny: the distribution fitting approach and the image analysis approach [@brown2010; @brown2005]. These are applied to architectural data series from well-preserved and documented archaeological samples within Linear Pottery and Trypillia contexts, as well as to synthetic data series.

For the distribution fitting approach, house-size distributions within settlements are modelled following a given procedure, and the retained model (the best fit) is interpreted in terms of social generating mechanisms. In particular, it is argued here that so-called power-law distributions reflect hierarchical structure, so that the identification of these within the studied samples may indicate the presence of some social hierarchisation process which warrants further interpretation (**rephrase?**).

With the image analysis approach, archaeological and synthetic settlement plans are analysed through the calculation of fractal dimension and lacunarity -- summary statistics which serve as quantifications of irregular spatial patterns or image textures. I argue here that geometrically irregular settlement plans are indicative of relative independence between households, while settlements that develop within geometrically regular grids indicate stronger overarching social structures, with a continuous range of possibilities in-between. The goal here is to test to which extent quantitative measures like fractal dimension and lacunarity may help differentiating between varying degrees of planning in prehistoric settlements.

one more, general - - assessing problems specific to doing this in archaeology.

## Some definitions

Just the minimum required to present the thesis in general terms.

-   Hierarchy @pumain2006

-   Inequality @kohler2018, @midlarsky1999

-   Social complexity @daems2021

-   Scale @west2017

## Main findings here?

Maybe leave to the end, and fill in, like abstract.

## Research ethics

-   Social complexity and evolution. Are less complex societies simple? Is that a bad thing?

-   Open science and open-source scripts

-   Terminology and spelling (British English for text. For geographical place names, Slovak special characters are kept as far as possible, even though it can be a pain in the xxx to render in Rmarkdown on Windows OS, and the 2010 Ukranian National transliteration system with only ASCII characters and no soft sign)

-   Abstracts in Slovak and Ukranian (and not only in Norwegian)

## Structure of the thesis

This thesis is structured as a monograph in four parts. In the first part the overall framework of the study is exposed, with the general introduction above, the overarching theoretical framework in Chapter \@ref(theory), and the background of the study material in Chapter \@ref(material). Parts II and III are devoted to each their methodological approach to the material: Part II to the study of hierarchy in size distributions, and Part III to the quantification of image textures. Each of these parts consists of three chapters, the first of which -- Chapters \@ref(house-sizes-theory) and \@ref(images-theory) -- exposes the theoretical and interpretative background of the applied methods and their relevance to Archaeology. The following chapters -- Chapters \@ref(methods-distfit) and \@ref(images-methods) -- detail the technical specifics of the two approaches and their implementation in this study, and the last chapters within these parts -- Chapters \@ref(results-distfit) and \@ref(images-results) -- provide the actual analyses and summaries of results. In Part IV, the findings are summarised and further discussed. Chapter \@ref(disc-results) gives an attempt of interpreting the results in the context of the culture-historical setting of the European Neolithic, while Chapter \@ref(disc-methods) reviews the possibilities and limitations of the fractal analysis framework in Archaeology. Concluding remarks and suggestions for further study are given in Chapter \@ref(conclusion).

Some readers might react to an apparent deviation from the academic tradition of devoting a separate chapter to research history. This is a deliberate choice, not to suggest that historiography is unimportant, but rather as a result of the fundamentally interdisciplinary scope of the study. In fact, there is very little extant history of applying fractal analysis in Archaeology -- the few studies that, to my knowledge, have been done in this direction are discussed primarily in Sections \@ref(distfit-archaeo) and \@ref(image-archaeo). Fractal analysis itself holds a research history of its own (see Section \@ref(fractals)), and so does the study of Linear Pottery (Section \@ref(lbk)) and Trypillia societies (Section \@ref(trypillia)), not to mention the general study of social complexity in Prehistory (mainly Section \@ref(complexity-archaeo)). In short, instead of trying to shoehorn these parallel histories into a clearly delimited but rather hybrid chapter, I have opted for what I believe to be a more useful approach, namely to fit them in more seamlessly where they belong, in the various associated theory and methods chapters.

An additional note should be made here regarding the writing style of the different parts and chapters. It is my belief that a major obstacle for fractal analysis methods to become more integrated into the standard tool kits of archaeological research, is the excessively technical nature of much of the associated literature. Archaeology as a discipline remains profoundly rooted in the Humanities, as seen in the inbuilt structure of teaching, research and funding institutions in most (at least European) countries. Fractal analysis is derived from pure mathematics, and most applications so far have been developed within the natural sciences (**ref necessary?**). Archaeologists who are trained within a humanistic scholarly tradition cannot be expected to hold a skill level of mathematics more advanced than what is achieved in high school, and code programming is hardly taught at all within the walls of Humanities faculties. Technical details regarding the methods and analyses applied in this thesis are therefore -- as much as possible -- limited to the devoted Chapters \@ref(methods-distfit) and \@ref(images-methods), and readers who are interested in these may also refer to the online code repository for more details and reproducibility **cite repository**. In the rest of the thesis I have opted for a more narrative approach, in an attempt to invite a somewhat larger audience of archaeologists into the fascinating complexities of fractals.

END Chapter

Here I skipped a line.\
Here I added a backslash.

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

Test: @morgan1965.

Test: @mandelbrot2021. jløkj fdsasdfadfasdfadf new text
