# (PART) Frameworks {.unnumbered}

# Introduction {#intro}

## Background of the study

-   Studying social hierarchy in Prehistory through fractal analysis of settlement plans

-   In Europe, the Neolithic is the long and messy transition period between mobile hunter-gatherer groups in the Palaeo- and Mesolithic, and the first city states in the Bronze (Aegean) and Iron Ages (Mediterranean and Central Europe)

-   Early farming economy, influence from the Near East.

-   Large variety in scale and content of archaeologically defined culture groups. Single farmsteads and small hamlets in many phases -- some with hardly any settlement evidence at all **examples mid-neo, use** @shennan2018. Other phases include exceptionally large settlements, probably hosting populations of several thousand inhabitants, like at Maidanetske in central Ukraine around 3.800 BCE (see \@ref(trypillia)). While in some settings, like in Linear Pottery society in much of continental Europe north of the Alps towards 5.100 BCE (see \@ref(lbk)), the dead were buried in simple pit graves, with very little distinction in treatment between individuals, in other phases some individuals were buried with tremendous amounts of precious goods like in Varna, Romania, or under colossal burial mounds like in Carnac, France, both in the mid-5th millennium [@shennan2018]**check ref**.

-   Seen at a very large scale -- across the continent and through the Holocene -- the development of society from small scale and relatively egalitarian towards large scale and more hierarchical seems evident (though not to everyone, see @graeber2021 and Section \@ref(soc-typ)). When we look more closely however, this evolution is anything but linear, as both population sizes and levels of hierarchical organisation seem to fluctuate considerably, sometimes over short time spans as from the Trypillia **C2 to D1??, check this**, when the so-called mega-sites are abandoned and their former inhabitants regroup into much smaller settlements during a transition of maybe only **check and ref** years.

-   In many cases, the level of social hierarchisation and complexity in a given Prehistoric society is very hard for researchers to evaluate, since many indicators of such structures are either lost from the archaeological record, or were never included in the first place [@perreault2019; Section \@ref(complexity-archaeo)]. Archaeological traces that are often interpreted as signs of social complexity and hierarchy may furthermore be deceiving. Seemingly monumental structures were in many cases built through small additions over centuries, rather than in one colossal construction campaign [**example, Carnac alignments? Danish megaliths?**]. In many megalithic burial contexts, it may be impossible to know how large a segment of the society that had access to such inhumations (**rephrase**).

-   New methods for investigating hierarchy: fractal analysis. Borrowed from other disciplines, not much tested in Archaeology. Example from human geography, use:[@jahanmiri2022; @lagarias2021; @themath2019; @tannier2005; @batty2005; @batty1994] (say what fractal analysis does, don't explain what it is here)(explain just the word fractal and cite section).

## Research question and objectives

The overall goal of the present study is to test and assess the utility of fractal analysis techniques as tools for studying hierarchical social organisation in prehistoric societies. Two methodological approaches are under special scrutiny: the distribution fitting approach and the image analysis approach [@brown2010; @brown2005]. These are applied to architectural data series from well-preserved and documented archaeological samples within neolithic Linear Pottery and Trypillia contexts, as well as to synthetic data series. This thesis is thus not to be considered a culture-historic study of Linear Pottery or Trypillia society, but mainly a methodological study. However, results from the proposed analyses of these case studies may also contribute, as side-effects, to their respective fields of research.

For the distribution fitting approach, house-size distributions within settlements are modelled following a given procedure, and the retained model (the best fit) is interpreted in terms of social generating mechanisms. In particular, it is argued here that so-called power-law distributions reflect hierarchical structure, so that the identification of these within the studied samples may indicate the presence of some social hierarchisation process which warrants further interpretation.

With the image analysis approach, archaeological and synthetic settlement plans are analysed through the calculation of fractal dimension and lacunarity -- summary statistics which serve as quantifications of irregular spatial patterns or image textures. I argue here that geometrically irregular settlement plans are indicative of relative independence between households, while settlements that develop within geometrically regular grids indicate stronger overarching social structures, with a continuous range of possibilities in-between. The goal here is to test to which extent quantitative measures like fractal dimension and lacunarity may help differentiating between varying degrees of planning in prehistoric settlements.

While both these methodological approaches are well developed and integrated to other disciplines, their usage in Archaeology have so far remained anecdotal [@diachenko2018]. A further overall goal of this thesis is to identify and explore possible limitations in the nature of archaeological data that may limit the applicability of fractal analysis methods within this discipline. For example, does fractal analysis of settlement plans require a data quality that would be practically unattainable in archaeological settings? But also inversely, as it is impossible to explore all potential applications within the framework of one doctoral thesis, suggestions for future research are provided in the last chapter.

## Defining hierarchies

The term *hierarchy* is central to the present study. Though commonly used in daily speech, defining the word is not as straight-forward as one might think, so some clarification on how it is understood here might be needed. In a volume dedicated to exploring the meanings and uses of hierarchy as a study object within a range of natural and social sciences, Denise Pumain provides a panorama of definition nuances, but also highlighting the characteristics that are commonly found in most cases [@pumain2006]. Among these characteristics are:

-   A pyramidal organisation of elements, ordered by a very unequal size distribution of a certain quality or variable, from a few large elements on top to many small elements at the bottom

-   When seen as a system, the whole is constituted of sub-systems, which are again constituted of sub-sub-systems, and so forth. These can either be ordered into clearly distinguished levels (stratified), or in other cases be scaled in a continuum (branched or tree structure)

-   In physical, biological and social hierarchical systems, the structure is often accompanied by a flow of energy, material, information or control in one or both directions between the top and bottom levels

Hierarchies are found in humanly constructed classification and taxonomical systems, where morphological distinctions are considered more important or fundamental at the higher end of the hierarchy, while being more detailed or specific at the lower end. Many hierarchical social systems, like religious (from Greek *hieros* -- sacred, and *archê* -- government), military or corporate organisations, include strongly reinforced regulations of subordination, which in modern society has led to somewhat negative connotations to both hyper-rational and despotic rule[@pumain2006:5-6]. While one prevalent explanation for the frequency of hierarchical structures in nature and society is indeed that they "represent the best solution for many optimisation problems" (*ibid*.:7), that does in no way mean they need to be consciously planned. On the contrary, in most cases hierarchies seem to emerge spontaneously, often from growth processes with systems splitting into sub-systems once they reach a certain critical size limit. There is also no compulsory link between social hierarchies and despotism, as it matters little to the overall structure whether the top element is elected for a limited period or born into an inherited leading position. More detail on how hierarchical structures emerge and how they can be described as fractals, is given in Chapter \@ref(theory). A further discussion on the specifics of social hierarchies is given in Chapter \@ref(house-sizes-theory), and on the differences between spontaneously emergent versus consciously planned structures in Chapter \@ref(images-theory).

A possible confusion with a somewhat different meaning of the term hierarchy should however be mentioned already here. If hierarchical structures are abundant in nature in both physical (inert) and biological systems, social hierarchies on the other hand -- understood as intra-species populations of individuals organised in pyramidal hierarchical, i.e. multi-level relationships to each other -- seems to be almost exclusively found among human groups. At the same time, a different type of hierarchy is frequently described by biologists which is common among animals, namely *dominance hierarchies*, also known as pecking order [@strauss2022]. These structures are hierarchical in the sense that there is difference in rank between group members, and they also seem to emerge spontaneously in the animal populations where they are described. But unlike social hierarchies, these are purely *linear*, in the sense that each group member is situated in rank above one part of the group and below the rest, so that the whole group forms a rank chain in the form A \> B \> C ... n. The rank of an individual will typically decide their access to food and reproduction relative to the other members, and may be settled and resettled in a number of ways depending on the species and population under study, but typically involving some level of violence or threat and subordination in face-to-face encounters [@strauss2022].

A classic example -- perhaps most of all in popular culture -- is dominance hierarchy among wolves, led by an alpha male [e.g. @cafazzo2016; @packard2003]. Though it has been much discussed whether or not this trope model actually fits wolves [see @mech1999; @muro2011], any reported dominance hierarchies among larger groups of wolves and stray dogs are linear rather than pyramidal, even when they are illustrated as pyramidal [e.g. Fig.1 in @rodríguez2017]. Similar social organisation systems are found among a wide variety of species -- mammals, birds, fish, particularly but not only among group-hunting carnivores -- and are generally interpreted as an evolutionary mechanism [@strauss2022, with references]. Cases of branching, multi-level social hierarchies among animals are on the other hand extremely rare, but have been reported to operate among hamadryas baboons in Ethiopia, with "clan leaders" forming relays of information flow and decision making between the "one-male units" within a total population ("band") of about 200 individuals [@schreier2009]. Eusocial insects like ants and wasps provide a more well-known example of hierarchical organisation among animals [@shimoji2022], also indicating -- if it should be necessary -- that it is not a matter of cognitive abilities or brain size, but rather of social function (e.g. building a hive or a village together) and population size above a certain threshold.

The reason to dwell upon this qualitative distinction between linear and pyramidal hierarchies, is that the prevalence of the former in nature is sometimes put forward as an argument for social hierarchies among hunter-gatherer groups in the Palaeo- and Mesolithic [find precise reference here --\> Graeber, if else strawman]. While it is a good point that there is no reason there couldn't be linear dominance hierarchies among small forager groups, and that such systems hardly can be described as egalitarian by those who live them, we cannot assume that pyramidal social hierarchies have always been part of human culture, but rather that they -- much like agriculture -- at some point were established **rewrite this**. HERE.

Bottom line hierarchy: pecking order/linear dominance hierarchy may be prevalent in any part of human Prehistory, but could also be very hard to study from material remains. Branching (i.e. fractal) hierarchies as well as social stratification should leave material traces, and are most probably related to population size (see Section \@ref(house-sizes-theory), and thus less likely to be found within pre-Neolithic societies. They should be regarded as historically situated phenomenons -- much like agriculture, pottery, writing or the steam engine -- justifying the search for and explanation of their possible origins [*contra* e.g. @graeber2021].

-   Inequality @kohler2018, @midlarsky1999, @price2010, @price1995 NA

## What is social hierarchy?

-   Political assumptions -- all hierarchical social structures are not despotic top-down rule. Democracies can also be very hierarchical. Matter of scale rather than political system @graeber2021. But, tendencies? Use @pumain2006, also @furholt2020. Use @redhead2022: status and leadership, multiple overlapping networks. Emergence of hierarchies: "Positions in the dominance hierarchy is determined by a combination of attributes of individuals, stochastic processes, and social context" @strauss2022.

-   Nested and non-nested social hierarchies, hierarchical hunter-gatherers? @hamilton2007, @whitridge2016 (or use in Chapter \@ref(disc-methods)?)

-   Biologically defined thresholds to group size? Dunbar's number and controversies, @dunbar2022, @west2023 (add published papers). Scalar stress @johnson1982, @alberti2014, @zhou2005. Also @carneiro1986

-   Temporal dynamics of social hierarchies: cyclicity [@peters2017], sawtooth waves [@scheidel2017], punctuated equilibrium [@gould2007; zimmermann?]. Archaeo. example South Sweden Neo/BA: @nordquist2001. Transitions villages to urban: @birch2014

<!-- -->

-   Tools for classifying societies, or evolutionary model? Discussion of @johnson1987, @testart2005, @service1971, @earle1997.

    -   Lineage and Chiefdom societies @earle2002, @sahlins2020

-   Anti-evolution critique in @yoffee1993, @yoffee2005, @fontijn2013, @kienlin2012, @lund2022, @furholt2020. Anarchistic critique/heterarchy: @crumley1995, @haude2019, @graeber2021

## Main findings here?

Maybe leave to the end, and fill in, like abstract.

## Research ethics

-   Social complexity and evolution. Are less complex societies simple? Is that a bad thing?

-   Open science and open-source scripts

-   Terminology and spelling (British English for text. For geographical place names, Slovak special characters are kept as far as possible, even though it can be a pain in the xxx to render in Rmarkdown on Windows OS, and the 2010 Ukranian National transliteration system with only ASCII characters and no soft sign)

-   Abstracts in Slovak and Ukranian (and not only in Norwegian)

## Structure of the thesis

This thesis is structured as a monograph in four parts. In the first part the overall framework of the study is exposed, with the general introduction above, the overarching theoretical framework in Chapter \@ref(theory), and the background of the study material in Chapter \@ref(material). Parts II and III are devoted to each their methodological approach to the material: Part II to the study of hierarchy in size distributions, and Part III to the quantification of image textures. Each of these parts consists of three chapters, the first of which -- Chapters \@ref(house-sizes-theory) and \@ref(images-theory) -- exposes the theoretical and interpretative background of the applied methods and their relevance to Archaeology. The following chapters -- Chapters \@ref(methods-distfit) and \@ref(images-methods) -- detail the technical specifics of the two approaches and their implementation in this study, and the last chapters within these parts -- Chapters \@ref(results-distfit) and \@ref(images-results) -- provide the actual analyses and summaries of results. In Part IV, the findings are summarised and further discussed. Chapter \@ref(disc-results) gives an attempt of interpreting the results in the context of the culture-historical setting of the European Neolithic, while Chapter \@ref(disc-methods) reviews the possibilities and limitations of the fractal analysis framework in Archaeology. Concluding remarks and suggestions for further study are given in Chapter \@ref(conclusion).

Some readers might react to an apparent deviation from the academic tradition of devoting a separate chapter to research history. This is a deliberate choice, not to suggest that historiography is unimportant, but rather as a result of the fundamentally interdisciplinary scope of the study. In fact, there is very little extant history of applying fractal analysis in Archaeology -- the few studies that, to my knowledge, have been done in this direction are discussed primarily in Sections \@ref(distfit-archaeo) and \@ref(image-archaeo). Fractal analysis itself holds a research history of its own (see Section \@ref(fractals)), and so does the study of Linear Pottery (Section \@ref(lbk)) and Trypillia societies (Section \@ref(trypillia)), not to mention the general study of social complexity in Prehistory (mainly Section \@ref(complexity-archaeo)). In short, instead of trying to shoehorn these parallel histories into a clearly delimited but rather hybrid chapter, I have opted for what I believe to be a more useful approach, namely to fit them in more seamlessly where they belong, in the various associated theory and methods chapters.

An additional note should be made here regarding the writing style of the different parts and chapters. It is my belief that a major obstacle for fractal analysis methods to become more integrated into the standard tool kits of archaeological research, is the excessively technical nature of much of the associated literature. Archaeology as a discipline remains profoundly rooted in the Humanities, as seen in the inbuilt structure of teaching, research and funding institutions in most (at least European) countries. Fractal analysis is derived from pure mathematics, and most applications so far have been developed within the natural sciences (**ref necessary?**). Archaeologists who are trained within a humanistic scholarly tradition cannot be expected to hold a skill level of mathematics more advanced than what is achieved in high school, and code programming is hardly taught at all within the walls of Humanities faculties. Technical details regarding the methods and analyses applied in this thesis are therefore -- as much as possible -- limited to the devoted Chapters \@ref(methods-distfit) and \@ref(images-methods), and readers who are interested in these may also refer to the online code repository for more details and reproducibility **cite repository**. In the rest of the thesis I have opted for a more narrative approach, in an attempt to invite a somewhat larger audience of archaeologists into the fascinating complexities of fractals.

END Chapter

Here I skipped a line.\
Here I added a backslash.

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

Test: @morgan1965.

Test: @mandelbrot2021. jløkj fdsasdfadfasdfadf new text
