# (PART) Frameworks {.unnumbered}

# Introduction

## Background of the study

## Research question and objectives

What are my goals for this project?

Two approaches: house-size distributions and settlement layouts

## Some definitions

Just the minimum required to present the thesis in general terms.

-   Hierarchy

-   Inequality

-   Social complexity

-   Scale

## Main findings here?

## Research ethics

-   Open science and open-source scripts

-   Terminology and spelling (British English for text. For geographical place names, Slovak special characters are kept as far as possible, even though it can be a pain in the xxx to render in Rmarkdown on Windows OS, and the 2010 Ukranian National transliteration system with only ASCII characters and no soft sign)

-   Abstracts in Slovak and Ukranian (and not only in Norwegian)

## Structure of the thesis

Interdisciplinary subject and no single history of research chapter.

Accessibility, non-specialised readers: technical details are, as far as possible, limited to the dedicated chapters (\@ref(part2)).

Describe the parts (including this one).

Text

Here I skipped a line.\
Here I added a backslash.

fdsasdfasdf You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
