# Theoretical framework: Scale, Complexity and Fractals {#theory}

## Micro-macro approaches in social and anthropological theory {#social-hierarchy}

### Sizes of human social groups

The scalar relations between overall social structures, the individual and everything in-between is a major theme in social theory, and any discussions on the nature of social hierarchies will quickly reach in to the very core of all disciplines that deal with human societies. Here I will not attempt to do justice to such a large topic of the social sciences and humanities, but rather outline the specific questions that seem the most relevant to this thesis and that have been the most lively discussed by archaeologists and prehistorians.

The political connotations to hierarchical social structures were already mentioned in the introduction, and will be evoked repeatedly throughout this thesis. The crucial analytical distinction between political system, economic inequality and hierarchical structure is often forgotten, leading to researchers with different interests talking past each other in fruitless debates. Top-down despotic political systems are not the only possible form of hierarchy, and inversely, a search for hierarchical structure in archaeological data is not a search for top-down despotic political systems [see @furholt2020].

A recurring theme in anthropological discussions related to hierarchy -- often termed social complexity -- is its purported correlation to group size. The notion of *scalar stress* was advanced in archaeology by @johnson1982, referring to how social groups that grow to a certain size will tend to reach a limit, beyond which they can only grow by splitting (fissioning) into smaller groups, as conflict levels rise and group solidarity and integration decrease [@alberti2014; @bandy2004; @bandy2008; @carneiro1986; @dubreuil2010; @drennan2008]. Evolutionary psychologists have proposed more causal explanations to why this happens, with the claim that the observed thresholds for maximum group sizes are largely biologically determined within species, and that subsequent fissions at higher hierarchical levels (as well as lower level groupings) tend to happen at multiples of the same threshold [@west2023; @zhou2005]. Most famously, Robin Dunbar proposed the "social brain hypothesis" in the 1990s, claiming that ideal group sizes among primate species were determined by the average size of their neocortex, with that of humans being around 150 individuals [see @dunbar2023 for a recent overview]. This idea has since been massively criticised for being too simplistic, according excessive importance to bio-genetic factors and ignoring culture [e.g. @lindenfors2021], while social media studies on very large data sets have given it more credit [e.g. @goncalves2011].

Some of the critique of what has become widely known as "Dunbar's number" seems to stem from a certain malaise related to the linking of social complexity and brain size, especially among anthropologists [e.g. @graeber2021]. Ironically, while social hierarchy often is given certain negative connotations like despotism, oppression and inequality, social complexity (which to a large extent is synonymous with hierarchy, see below) is often given a positive connotation in the sense that it is indicative of highly developed systems. Such connotations are indicative of the common confusion of the terms complex and complicated. A central tenet to complexity theory is that complex phenomena, like social hierarchies, are generally not consciously planned -- they *emerge*, through repeated interactions between smaller components who obey more or less simple rules. In other words, claiming that social complexity or hierarchy emerges in large human populations, and therefore generally only from late in prehistory and onwards [e.g @ross2017; @smith2011], is not to say that these complex societies are in any way better or smarter than less complex (often forager) societies -- they are simply different. Anthropologists who deal with comparative big history routinely express the need to point out that foragers also can be highly complex [e.g. @graeber2021; @testart2012], which is of course true in many ways, but somewhat beside the point when regarding hierarchical organisation in particular. There has never been such a thing as forager emperors [though @graeber2021 do discuss some apparent exceptions].

As a small side note, some specific types of social complexity are particularly relevant for the study of forager societies. One is the phenomenon of nested hierarchies, a trait probably common to all human societies, where any single person belongs to a variety of social groups of different scales, which are made relevant in different social settings. A person can belong to a family, an extended family, a village community, an ethnic minority, a nation state, even have a continental or human identity along with many other identities at the same time. The total of all these identities form a nested hierarchy, and is indeed fractal -- there are a few large ones and many small ones -- but it is not the same as a hierarchically organised society. Nested social hierarchies among foragers is an interesting topic [@hamilton2007; @whitridge2016], which is in no way in contradiction to the general framework of this thesis. Another term which has been much debated for many of the same reasons, is that of *complex hunter-gatherers* [@arnold1996], often claimed to invalidate general narratives of increasing complexity over the Holocene and its link to agriculture. These are societies now known from many parts of the world, but originally associated with north-west coast Native American groups known for their sedentism, complex art forms, extensive food storage and competitive feasting organised by clan or house leaders (see below). To which extent they constitute exceptions of or confirmations to rules of how foragers are organised is a different debate, but it should in any case be kept in mind that they are often associated with specific eco-zones with abundant marine resources which allow them to sustain relatively large and dense populations.

Though a very general tendency of increasing levels of social hierarchy or complexity is recognised by most big history researchers, much interest has been accorded to deviations from this rule, to the extent that many authors consider such deviations to be systematically inherent in how human societies evolve over time, either in cycles [e.g. @peters2017; @zimmermann2012a; @gronenborn2014] or in more abrupt movements [@scheidel2017; @heitz2021; @gould2007]. In any case, it is today widely recognised that societies do not simply grow towards ever more complexity in a linear manner [@graeber2021].

### Social evolution and its critique

Another closely related approach to social hierarchies which has been subject to much debate in archaeology, is that of societal classifications, usually grounded in social evolutionary theory [@johnson1987; @service1971; @testart2005]. In general, such works have aimed to define typologies of stages of social evolution based on comparative anthropology, more or less explicitly seeing current small-scale societies as remnants of society forms that would have been representative of earlier periods in human history and prehistory [@dubreuil2010; @ragan2017]. The perhaps most widely (by archaeologists) cited work was done by @johnson1987, who proposed a scheme of classifying small-scale societies into three main classes: those defined by the family group (without or with domestication), the local group (acephalous or "Big Man" collectivities) or regional polities. The latter class was further subdivided into chiefdoms, early or "archaic" states and nation-states with peasant economy. The focus of studies within this tradition has often been on types of political leadership, and explaining their functioning and temporal dynamics [@earle1997; @earle2002]. As an example, the leader type known as the Big Man -- modelled on ethnographic accounts from the New Guinea Highlands -- is recognised as one who bases his (male) power over the local, usually single village, community on economic surplus and prestige, but not heredity like higher level chiefs. Big Man societies are thus thought to be politically more unstable, since the leader can be challenged, and in any case at his death someone else will take the opportunity to fill his place [@johnson1987, pp. 160-93]. Chiefs on the other hand, build their power on inheritance, and high level chiefs ruling over local chiefs are known to rule "complex chiefdoms". As such, within the social evolutionary framework, the type of society is closely linked to population and scale.

While such classifications are widely used and referred to by archaeologists specialising in prehistory (perhaps especially Neolithic and Bronze Age societies), they have been subject to substantial criticisms. Specific categories have been shown to gloss over important differences between the societies they are supposed to characterise, thus making us blind to variability [@yoffee1993; @pauketat2007; @fontijn2013; @kienlin2012; @yoffee2005]. They have been criticised for promoting simplistic views on political power [@lund2022; @furholt2020]. Anarchistic theory has shown that what authors like @johnson1987 label as acephalous systems are not limited to small bands of mobile foragers, and systematic social difference does not necessarily involve vertical hierarchies [@crumley1995; @haude2019; @graeber2021; @angelbeck2022]. Furthermore, the assumption that Palaeolithic foragers were necessarily egalitarian has been much debated, with social inequality in the so-called "princely graves" of the Gravettian being a recurrent theme [e.g. @graeber2021; @honoré2022; @testart2012; @testart1982]. A somewhat ironic aspect of many of these discussions is that critics who point to the simplistic nature of societal classifications, in the end tend to replace or add to the categories themselves. In fact, the now classical works of mid to late 20^th^ century neo-evolutionary authors were originally also conceived as critiques of 19^th^ century evolutionary or Marxist models [e.g. @engels1902] perceived as too simplistic and not founded on empirical observations.

One such additional category of social structure, which has had a certain success in archaeology, was described by Claude Lévi-Strauss as *house societies* [@levi-strauss1982; @joyce2000; @beck2007]. Here a house is a social unit, often centred around a material estate but also involving titles, heirlooms, land ownership, rights to hunting grounds etc. and where membership is not determined from genealogy in a systematic manner, as is the case with more regularly structured lineage or clan societies. House membership may be gained through more competitive social action, privileging those that possess the resources to engage in activities such as gift exchange and *potlatch*-style feasting [see @hayden2014]. This opens up for more complex configurations, and house membership will often entail a certain level of prestige. Inheritance may furthermore follow (male) descent or (female) affinity -- that is from grand-father to grand-son via mother -- in a pragmatic way depending on which option is in the best interest of the house, as long as it can be justified in more or less precise kinship-related terms [@sahlins2013; @ensor2011]. A segment of society of varying size will not afford to partake in this competition, and as a result, house societies can be viewed as being in a somewhat unstable transition state between lineage and class societies. Lévi-Strauss originally pinned the term on Kwakiutl society in the American north-west coast (as described by Franz Boas in the late 19^th^ century), but argued for its generality by associating it with the feudal system of medieval Europe. The concept has since been applied to a wide range of cases from ethnography (especially in the South-East Asian and Pacific regions), as well as to prehistoric contexts [e.g. @kuijt2018].

The subtext of debates regarding social evolution, scale, and types of societies -- as I see it -- is disagreement regarding what is most important, between shared versus unique traits, or overall trends versus variation. Broad evolutionary models are systematically criticised for being too simplistic, while adding more and more nuance can make one loose sight of similarities and trends that are also real enough. In the long run, such debates inevitably take on the characteristics of false dichotomies -- both points of view are correct, but largely unable to see the value in other approaches, which is why the debates do not advance much but rather seem to go in loops.

My stance in this thesis is that in order to better apprehend the complex relationships between different scales of society, and individual agency versus structure in particular, other frameworks are needed. A number of valuable approaches have indeed been proposed within social and anthropological theory precisely as attempts of bridging the gap in different ways between the micro and macro scales of society, but which have been less regularly referred to in archaeology (note that this thesis, while written at a Scandinavian university, deals with case studies and associated academic traditions in continental Europe, where neo-evolutionism is still largely dominant in late prehistoric research). These include, but are not limited to, structuralism [@levi-strauss1969], world systems [@wallerstein2011; @braudel1976], structuration theory [@giddens1984], actor-network theory [@latour2005] and assemblage theory [@delanda2006]. While there is certainly not enough space here to present all of these theoretical approaches and the differences between them in detail, it can be noted that they are all mostly associated with qualitative methods. It is my conviction that the massive amounts of archaeological data which has become available in recent years [@bevan2015], also calls for the use of quantitative methods and associated theory in research on social complexity and hierarchical structures across scales. The most well-suited framework for this purpose should be dynamical systems theory, which covers fractal analysis among other methods, and which -- to the extent it is applied in the social sciences and humanities -- is often labelled as complexity theory, and has strong parallels to qualitative approaches like assemblage theory. Exploring how these very separate theoretical traditions approach many of the same phenomena in the real world, would be the subject of entire theses in itself.

## Complexity Theory / Dynamical Systems Theory {#complexity}

The study of complex dynamical systems derives from physics and the natural sciences, where such systems are characterised by multiple actors or agents -- living or inert -- that interact with each other following certain rules, and from which new behaviours emerge at larger scales. Such *emergence* is an overall pattern, behaviour or structure that is described as more than the sum of its parts. Common examples are galaxies, ant colonies, river systems, weather systems, flocks of birds, and so on. @delanda2006, who in my view describes exactly such phenomena within human societies and with humanist terminology, argues that at certain points the system becomes an agent of its own. Dynamical systems theory uses a range of concepts for describing and explaining such systems, and it is widely recognised that most of them also apply to humans as well (there is no reason they should not). However, they have only to a very limited degree been integrated in the humanities, including archaeology. For a technical introduction, see @devaney2020. A more lay-person introduction is given by @west2017. Applications in archaeology include network science, settlement scaling, cultural evolution and agent-based modelling [@daems2021; @baden2016; @ross2017; @smith2011; @bentley2008]. Some central concepts in dynamical systems theory, including chaos, feedback loops, criticality and scaling, are discussed in some more detail in Chapter \@ref(house-sizes-theory).

Social complexity is more than social hierarchy, and all societies -- whether they are classified as simple or complex within a social evolutionary framework -- can arguably be studied as complex systems, since they always consist of various sub-systems and populations of individuals acting and interacting in a variety of ways [@daems2021, p. 6]. The sense of social complexity that is traditionally understood in archaeology, derived from social evolution theory, is somewhat narrower and relates more to the specific characteristics of organisational scale (i.e. hierarchy) than to complexity *per se*. Studying the scale of social organisation from a complex/dynamical systems perspective rather than from a social evolutionist one holds several advantages:

-   The switch from discrete evolutionary stages to continuous spectra allows for more nuanced evaluation of the society in question, avoiding false binaries like simple-complex

-   The complex systems approach arguably has a stronger explanatory power than the more classificatory social evolution schemes

-   Complexity theory offers a better alternative for ethical reasons, as it avoids the underlying colonial and eurocentric connotations associated with classifying societies into simple and complex

As a further elaboration on the latter point, an analysis of a society from a complexity theoretical viewpoint is not a matter of establishing whether or not the society can be characterised as a complex system. When the scale of social organisation is the object of study -- as in the present thesis -- the word hierarchy is both more accurate and, if not neutral, at least more balanced than social complexity, since it is not obvious whether or not it is a good thing for a society to be characterised as hierarchical. The study of the dynamics of social hierarchy over time is thus not a story of progress, as the 19^th^ century studies of social evolution too often were.

### Fractals and Fractal Analysis

Fractal analysis relates to complex systems/dynamical systems in that it is the methodological framework which allows for studying the complex patterns that they tend to generate [@mandelbrot2021; @falconer2013].

The term *fractal* was pinned by the French-American mathematician Benoît Mandelbrot (1924-2010), from the latin word *fractus* meaning broken or irregular, to describe patterns that because of their apparent limitless complexity defied concise description within the frameworks of Euclidean geometry and classical calculus [@falconer2013, pp. 116-20]. Such patterns -- both theoretical and empirical -- had been described and analysed by mathematicians and researchers within other disciplines since the end of the 19^th^ century, but were mostly regarded as curiosities and exceptions, and Mandelbrot was the first to link all these previous studies within a unified theoretical framework [@mandelbrot1977; @mandelbrot2021].

In one influential paper, drawing on previous work by mathematician Lewis Fry Richardson, @mandelbrot1967 argued that a rugged linear feature like a coastline could not be fully described through traditional geometry with a set of line segments, since this would result in a curve of infinite complexity. More importantly, he showed that the traditional measure of lines -- the length -- will inevitably depend on the scale of observation when applied to a coastline. If measured in kilometres, a coastline will always appear shorter in total length, than if it is measured in metres, since smaller bays and inlets can then also be accounted for. But this phenomenon continues seemingly without limit, since the same coastline measured in centimetres will appear much longer, and in millimetres far longer again, and so on (Figure \@ref(fig:02-koch)). Length as a measure of rugged linear features thus seems inadequate, which may become a problem in practical settings when comparing coastlines between countries that operate with different measurement units and procedures. The same problem occurs when describing irregular patterns in the plane (like island or continent outlines) with area or in three-dimensional space (like clouds or galaxies) with volume. As a solution, Mandelbrot proposed the use of the *fractal dimension* as a descriptive tool for characterising such patterns (see Chapter \@ref(images-theory) for definition).

(ref:02-koch) Construction of mathematical fractals, illustrated by the von Koch curve, with its generator (top), iterator (middle) and limit set as iterations approach infinity (bottom). For each segment, the middle third is replaced by two segments forming sides of an equilateral triangle. In theory, the von Koch curve has infinite length, and is nowhere differentiable (has no tangent). All figures by author unless stated otherwise

```{r 02-koch, fig.cap="(ref:02-koch)", fig.scap="The von Koch curve"}
load("Results/fig02_koch_ill.RData")
fig02_koch
```

Mandelbrot furthermore defined fractals as patterns, shapes or structures that exhibit *self-similarity* and *scale invariance*, meaning that their overall features consist of copies of themselves in ever smaller scales. At smaller scales, these copies also become ever more numerous, so that their size distribution follow a *power law* (see Chapter \@ref(house-sizes-theory)). In this sense, they are geometrical expressions of hierarchies.

Fractal shapes are found almost everywhere in nature -- plants, geological features, clouds, river systems and galaxies, inside our bodies in our nervous systems, lungs and blood vessels [@barnsley1993] -- but also in many humanly constructed features, like cityscapes, settlement systems, road and cable networks, waste deposits and so on, and can be used to model more abstract structures like demographic patterns and decisional hierarchies [@batty2005; @diachenko2018]. Many processes and mechanisms have been shown to produce fractals, and they often include dynamical growth, emergence and chaos (these processes are further discussed in the following chapters). The jargon in fractal analysis literature often borrows terminology from biology and fluid dynamics to allude to fractals in nature, like cascading bifurcations and confluences (indicating splitting or merging), tree structure or arborescence for branching, and turbulence for chaotic and irregular movement or behaviour (Figure \@ref(fig:02-trees)).

(ref:02-trees) Fractal shapes are ubiquitous in nature -- here represented by branching trees

```{r 02-trees, fig.cap="(ref:02-trees)", fig.scap="Trees as example of fractal shapes in nature"}
include_graphics("Results/fig02_trees.pdf")
```

Some common features are seen in most settings in nature and society where fractal patterns are found: the involved processes are un-planned, they emerge spontaneously from many iterations of simple rules that are followed by many smaller parts that each follow independent trajectories. These features have some important implications. While the small independent agents (e.g. water molecules) obey simple deterministic rules, outcomes are largely unpredictable. If a raindrop falls from a cloud over the Alps -- at a critical point -- infinitely small variations in wind direction and intensity can decide if its final destination will be in the Adriatic or the North Sea. Such systems are both deterministic *and* unpredictable -- they are *chaotic*. The concept of *critical* or *tipping point* is expressed in daily speech with expressions like "the straw that breaks the camel's back". Small additions of randomness to entirely regular and mathematically defined fractals can generate highly complex shapes, effectively imitating shapes we see around us in nature (Figure \@ref(fig:02-fractals)).

(ref:02-fractals) Examples of fractal shapes: the Sierpinski triangle (left), a regularly self-similar spiral (middle), and interference between two self-similar spirals starting from different positions, generating a turbulent pattern (right). These shapes -- as well as the von Koch curve above -- were generated by chaos games, a type of Markov Chain Monte Carlo algorithm [@falconer2013, pp. 17-29], and plotted in R using *ggplot* functions [@wickham2016]. While highly complex, the code for generating them is remarkably simple (see online repository for details)

```{r 02-fractals, fig.cap="(ref:02-fractals)", fig.scap="Examples of fractal shapes"}
load("Results/fig02_fractals.RData")
fig02_fractals
```

In addition to being embedded in space, fractal structures can be seen in more abstract levels:

-   Time series: earthquakes, finance [@mandelbrot1997]

-   Networks and abstract hierarchies: organisations and companies, income distributions, word counts, 1/f or pink noise, the World Wide Web (see Chapter \@ref(house-sizes-theory) for details)

-   Pure mathematics: Julia and Mandelbrot sets, strange attractors [@falconer2013]

Fractal analysis is developed for studying irregular phenomena (methods are described in more detail Chapters \@ref(methods-distfit) and \@ref(images-methods)), and thus constitutes a tool box for quantitative empirical research. After a wave of apparent over-optimism to the ubiquitous nature of fractals and the possibilities offered by their analysis in the 1990s and early 2000s -- as seen in publication titles like *Fractals everywhere* [@barnsley1993] and *How Nature Works* [@bak2013] -- it is now more widely recognised that this framework, like anything else, has its limits. Searching for these limits when methods are applied to archaeological settlement data is the overall topic of this thesis.
