# Material and data: Neolithic settlement plans {#material}

## Studying social hierarchy in archaeology and prehistory {#complexity-archaeo}

Grave goods

Example of Varna: Lichardus (1991; also Gimbutas), as discussed in @kadrow2013, proposed that the social inequality observed at Varna in the mid-5th millennium, was already then adopted from pastoral North Pontic steppe cultures (*check this and delete if unsure*); cf. @chapman2006 -- three social levels interpreted from grave goods, but not clear if these are really discrete or constructed on a continuum. The term *fractal* is here used purely as a metaphor for complex personhood structures, rather in opposition to the described social hierarchy. Biehl and Marciniak in Diehl 2000: no hierarchy in Varna. Response by Hayden 200?

Burial monuments

Architecture @rathje1982 -- social stratification (qualitatively) through architecture and burials (and life expectancy). Short but to-the-point article.

The denominator problem

The use of ethnography

Other approaches (osteological, isotopes, craft specialisation **refs**)

This project: house-size distributions and settlement layouts (details in subsequent chapters), just very short argumentation.

Comparative approach: Neolithic technology (not bronze axes), wood and wattle-and-daub architecture, (near) complete settlement plans/extensive documentation: Trypillia and Linear Pottery

For this project I opted for the use of houses and built environments as proxy for social hierarchy. In this way I hope to largely avoid the denominator problem associated with burials. While it is true that for many archaeological culture groups in late prehistory habitats are poorly preserved and hard to discover, leaving us still with a limited understanding of them (the case in many Michelsberg, Corded Ware and Bell Beaker groups, only to mention a few **check**), in groups where habitats are well preserved, there is little reason to suspect that the available record would not cover the whole range of social statuses if these societies were hierarchical. Unlike burials, every individual in a sedentary society -- with few exceptions like homeless persons in more recent urban contexts -- will normally have at least one fixed place to stay overnight, and these homes will in most cases be constructed within the same fundamental framework of techniques and building materials, depending more on culture specific traditions and environmental factors than on social status (\#**citation?**). As an example, in a society where mudbrick is the main building material, like in the Neolithic Near East and Anatolia, nearly all constructions are made in mudbrick, regardless of the social status of the inhabitants. In continental Europe north of the Alps, wattle-and-daub construction was the almost exclusive building technique for any architectural feature from the early Neolithic until the Roman conquest, and well into the Middle Ages north of the *limes*. One can of course enumerate exceptions, but more importantly houses are in any stratified society also a marker of social status, which can be exhibited in a range of ways, from decorations, use of more precious raw materials as well as size. That is precisely the reason for using houses as a proxy for social status and hierarchy in archaeological settings. But the point here is that there should be little taphonomic differentiation between groups of high and low status within a given archaeological context, at least in prehistory, and at least not as much as can be expected for burials, meaning that we can expect to find samples of houses that are representative of the social structure of the archaeological culture in question. On sites where there is taphonomical loss of architectural structures, as long as the overall building tradition is homogeneous, there is no reason this loss should affect one segment of the society more or less than others.

Some caveats do remain, however, for the use of houses as proxy of social status. Firstly, there may be a documentation bias favouring larger houses, since they may be easier to discover both during excavation and in remote-sensing surveys (\#**citation?**). In samples from originally very skewed house-size distributions, there may also be a further taphonomic bias towards large houses, since smaller houses -- being far more numerous -- are statistically in greater risk of being affected by post-depositional disturbances. Both of these biases are hard to evaluate empirically, though computer modelling could potentially give indications of their importance. This, however, is not within the scope of the present study.

A second, and maybe more important issue, is that of the contemporaneity of houses. When the goal is to investigate the social structure of a settlement as reflected in its architecture -- be it through the size distribution of buildings or their spatial layout -- all the analysed features should ideally have been in use at the same point in time. This is however very hard to achieve in most archaeological settings, and many researchers choose to either ignore the issue, or to accept a temporal resolution that is far wider than what their research questions should logically allow for [@perreault2019]. One way of limiting this problem is to select study samples with little to no stratigraphic overlap, which might indicate short occupation span, though as shown below this indicator can also be deceiving. For both study areas selected for this thesis -- the Linear Pottery in south-west Slovakia and the B2/C1 Trypillia of central Ukraine -- settlement plans show very little overlap between houses, even though some of them probably developed over more than three centuries, as shown by radiocarbon dates and modelling (see below). Such settlement plans may be impossible to differentiate into separate coeval time samples without precise dating of construction and abandon of every individual house, counting in the thousands on the Trypillia mega-sites (an alternative method is presented only for the Linear Pottery settlements in Section \@ref(lbk)). On the other hand, the fact that there is so little overlap between houses despite temporal differences clearly illustrates how these settlement plans emerged over time, not as a *tabula rasa* in each generation but rather with new constructions respecting the location and orientation of older ones long after abandonment. Though such practices are indeed interesting, it is not at all obvious, however, to which extent they may reflect or even relate to social factors such as hierarchy. With settlement types with much higher degree of stratigraphic overlap, like the tell sites of the Balkan and Near Eastern Neolithic traditions, it may be easier to distinguish more or less coeval occupation phases, but they are again harder to document extensively -- because of the high density of archaeological finds and features, excavation surfaces typically cover only very small portions of the settlement, while remote sensing performs less well and not allowing for temporal disentanglement of constructed features (\#**citation**). In any case, the issue of temporal resolution of the data and its influence on analytical results is crucial, and will be discussed repeatedly throughout this thesis, with a summary in Chapter \@ref(disc-methods).

Comparative technology

Meaning of "house"

In the following I propose a very brief overview of the general archaeological contexts of the two cases analysed in this thesis, namely the Linear Pottery and the Trypillia cultures. Both have been extensively studied and enjoy long and rich research histories, and any attempt at a comprehensive overview here would be futile. In stead, I focus on the aspects that are the most relevant for the present study -- in particular of architecture, settlement layout and social organisation -- and address the interested readers to more complete reviews. In the final sections of this chapter I provide some discussion on the nature and limits of the data material used here, which consists mainly of geomagnetic imagery with expert interpretation of archaeological features, as well as synthetic data for simulations.

## The Linear Pottery culture complex {#lbk}

The Linear Pottery culture complex -- often referred to as the LBK from the German designation *Linearbandkeramik* -- represents the initial transition to agriculture and permanent settlements in continental Europe north of the Alps. It is recognised through a number of defining traits, seen primarily through pottery style (the name alludes to the commonly seen decorative patterns of incised linear bands on pottery vessels), but also through architecture, subsistence economy, burial practices and various aspects of technology, showing a remarkable coherence across a very wide geographical range between northern France and western Ukraine [@bickle2013a; @shennan2018]. As hamlet and village-dwelling farmers using already fully domesticated plant and animal species, including wheat and barley which do not occur in wild varieties in central Europe, the Linear Pottery phenomenon stands archaeologically in stark contrast to the preceding Mesolithic foragers in the same regions. It was therefore posited early on that it was driven primarily by migration from south-east Europe via the Carpathian Basin along the Danube [e.g. @childe1929; @whittle2022]. This assumption has later been confirmed by a number of proxies, including pottery seriations and ^14^C dating showing that the central characteristics (pottery, burials, architecture) were formed in the Lake Balaton region/Transdanubia (Hungary) around 5600-5500 BCE [@bánffy2020], before spreading progressively to Slovakia, Lower Austria, Moravia (Czech Republic), and from there westwards through Bavaria, to the Rhine and Seine basins, as well as north to Saxony and eastwards along the northern slopes of the Tatras and Carpathians through southern Poland to western Ukraine and Moldova [@saile2020; @bickle2013a]. More recently, genetic evidence has shown that the populations of Linear Pottery settlements were closely related and were almost exclusively of south-east European, and ultimately to a large extent Near-Eastern/Caucasian ancestry -- there was in other words very little if any genetic mixing with local forager groups [@szecsenyi-nagy2015]. This also corresponds to the origin of the domesticated animals, some of which could in theory have been domesticated locally (cattle from aurochs and pig from wild boar).

More subtle evidence from Linear Pottery finds tends to link them, to varying degrees depending on the regions, to local forager populations, and there has been considerable discussions over the years regarding the exact nature of such farmer-forager contacts. These traces include the persistence of some Mesolithic bead types and associated material use throughout central Europe [@rigaud2015], regular inclusions of the very contrasting Limbourg and La Hoguette potteries in the westernmost settlements, or similarities in lithic production and technology [see @bickle2013a, pp. 5-6 for a more detailed overview]. However all in all the Linear Pottery culture is today seen primarily as the archaeological remains of people who spread across the fertile loess soils along river valleys in central continental Europe, driven by demic diffusion -- i.e. gradual rather than abrupt spread (though rapidly in archaeological time scales, and in leapfrog steps between favourable areas), with new homesteads established a few kilometres further up or down the valley for every generation, due to population pressure [@dubouloz2008; @shennan2018; @bocquet-appel2008; @ammerman1973], possibly also involving competition for the best cultivation spots [@shennan2008a] and social prestige in establishing new settlements [@frirdich2005]. While slash-and-burn cultivation was earlier assumed to be an explanatory factor for the Linear Pottery expansion (e.g. by Gordon Childe), more recent archaeobotanical studies have pointed to "intensive garden cultivation" as a main cultivation strategy, involving long-term field maintenance spanning several generations and considerable work investment, possibly motivating claims to land ownership and inheritance [@bogaard2012].

More cultural variability is seen towards the end of the Linear Pottery time period, for which a number of regional sub-groups have been defined, primarily though not exclusively based on diversified pottery styles (Notenkopf/Želiezovce in the south-east, Šarka in the north-east, Blicquy/Villeneuve-Saint-Germain in the west), though often with strong continuity in settlement patterns with the more "classic" earlier phases. In most areas the end of the last phase, dated approximately to 5050-4950 depending on regional variations, is associated with a more abrupt rupture in material culture, followed after some time by new groups with more markedly different material culture and/or settlement systems (Lengyel, Hinkelstein, Großgartach, Cerny) in some cases after an apparent depopulation in the region [@peters2017; @gronenborn2014; @riedhammer2018]. Furthermore, over the last 30 years or so, there has been a steady increase in evidence of considerable social tensions leading up to this apparent disintegration of the Linear Pottery culture. Large enclosures, possibly for defensive purpose, were constructed around villages across the culture zone, and a number of mass graves within the enclosure ditches dating from this phase have been identified. At Herxheim (near Karlsruhe in Rheinland) @boulestin2009 have reported on the remains of up to 1000 individuals of various origins (judging from the accompanying ceramics), having been violently dismembered and in many cases showing clear traces of cannibalism. At Asparn/Schletz north of Vienna, a smaller massacre deposit containing 67 individuals is interpreted as the remains of a single-event slaughter of the entire village. The age-sex ratios of the deceased are representative of expected values for a Neolithic village community, apart from the absence of young adult females, leading @teschler-nicola2012 to the conclusion of mass abduction by the attackers. Other sites from this phase which include human remains disposed *en masse* are known from Germany, Hungary and France, but not all show clear signs of violence [e.g. @meyer2014]. Yet an example is the large settlement of Vráble in Slovakia, presented below, of which the human remains are currently under study [@muller-scheessel2021; @furholt2020b]. Preliminary results point to a diversity in treatment of the deceased, and again clear traces of direct peri-mortal violence are largely lacking, adding nuance to earlier claims of societal crisis and warfare leading to the end of the Linear Pottery culture [e.g. @faruggia2002]. Nevertheless, the concentration of these events -- however diverse -- towards the end of the period, combined with their wide geographical distribution, has reanimated earlier debates regarding the internal dynamics of Linear Pottery society, and the possibility that increased social inequality lead to the observed social instability and decline [@augereau2021; @jeunesse2022].

The Linear Pottery settlements sampled for analysis in this thesis are located within the small river valley of the Žitava in south-west Slovakia (Figure \@ref(fig:03-zitava)). #Finish this descriptive sentence.

### Sampled Linear Pottery settlements: the Žitava valley

(ref:03-zitava) Location of the Linear Pottery settlements analysed in this thesis, in the river valley of the Žitava, a small left tributary (via the Nitra and Váh rivers) of the Danube in south-west Slovakia. The large Linear Pottery burial site at Nitra is located about 22 km to the west of Vráble. Terrain model data in public domain, courtesy of the U.S Geological Survey (<https://earthexplorer.usgs.gov>)

```{r 03-zitava, fig.cap="(ref:03-zitava)", fig.scap="Map with location of analysed Linear Pottery sites"}
include_graphics("Results/fig03_zitava.pdf")
```

### Linear Pottery houses, burials and society: current perspectives

-   origin of the longhouse: @bánffy2020, @coudart2015, @last2015

-   house typology, construction, and use: @modderman1970, @coudart1998

Number of coeval houses in LBK villages, evaluated to a mean of 5, sd. 3 [@coudart1998, p. 91]. Furthermore, based on ethnographic analogies she estimated a typical Linear Pottery village of 3 to 8 houses (of which 1 or 2 longer than 22 m), to include on average approximately 150-260 inhabitants, though fewer (up to around 80 people) for the smallest hamlets of 3 houses.

-   Settlement layout: Yard model @zimmermann2012, @petrasch2012. house-row model: @ruck2012. Middle-way: @link2012, and regional variability: @lenneis2012.

    -   The yard model implies some degree of land inheritance. House generations

-   Settlement structure: Central places (large+enclosures), peripheral smaller. Archetype: Aldenhoven plateau in Rhineland [@classen2005]

    -   Systems within secondary river valleys, e.g. Stephansposching and other sites on the left bank of the Danube in Lower Bavaria [@pechtl2012]. Also in Lower Austria [@pieler2012]. The Aisne valley [@dubouloz2012] Similarly the Zitava valley. Interpretation of this apparent settlement hierarchy varies: @vandevelde1990

    -   Burials: @jeunesse2022, check @gaydarska2023 and @augereau2021. @frirdich2005

        -   Nitra cemetery: @pavuk1972. 74 graves. 9 with polished stone adzes. Interpreted as egalitarian, though jeunesse -- old men.

        -   Patrilocality in LBK based on Sr isotopes: Bickle/Whittle. Suggested by genetics: @rasteiro2013; @szecsenyi-nagy2015; @rasteiro2012, house-size distributions: @schiesberg2016. More variability (including polygyny, abduction of young women and non-inhumation burials): @hrncir2020a.

        -   Patriarchal society in Central Europe first appears in early Linear Pottery communities at Nitra, @augereau2021, p. 221. Man-with-adze as the first arrival to the (best) spot [the "ideal despotic distribution" model proposed by @shennan2008a]

        -   Enclosures and warfare as evidence of integrated village communities and possibly leadership [kerig 2003]

While it was already recognised by Pavúk that polished stone adzes (Ger. *Schuhleistenkeil* or *Dechsel*) were only found in male graves in the Nitra cemetery, and more so predominantly in the richest graves and among older males [-@pavuk1972, p. 55-6], he concluded that this finds category should not be interpreted as anything more than a utilitarian tool for woodworking.

### Time samples of the Vráble settlement (moved from Chap. 6, review to check it fits)

Separating an archaeological settlement plan into samples with temporally coeval structures only is a demanding exercise, and more so the larger the settlement. The best and most reliable method is to date every single structure in the settlement, but this is usually impossible both in theory, since not all structures yield datable material, and in practice when considering the costs involved. The more realistic approach is to sample a smaller amount of structures throughout the settlement for dating and modelling, and then extrapolate the results to get a more or less rough overview of how much, proportionally, of the site was occupied at the same time. This has been done for the Trypillia mega-sites, indicating that the maxima of coeval habitation would be at approximately 33% of houses at Nebelivka [@millard2020, pp. 253-6; @muller2022] and 52% at Maidanetske [@ohlrau2020, pp. 233-5; @ohlrau2022, pp. 86-8]. Even though some general tendencies of the spatio-temporal development of these settlement plans have been proposed [@shatilo2021, p. 247; @muller2022, pp. 218-9], it is for now impossible to extract precise temporal samples of coeval house-size distributions here. This also goes for the image analysis presented in Chapter \@ref(images-results) [see also @bruvollforthcoming]. The smaller sites of Moshuriv and Talne 3 are only dated through relative chronology of surface finds and thus attributed as a whole to their chronological phase, though their small sizes and ordered layout would suggest that most houses could be coeval [@ohlrau2020, pp. 241-4].

house orientation: "origin" hypothesis: @bradley2001

For houses in Linear Pottery settlements, a special dating proxy was recently developed, based on the observation that within this cultural context houses were constructed to be parallel to pre-existing ones, but with a slow and gradual counter-clockwise shift in orientation of approximately 1.3 degrees per decade [@muller-scheessel2020]. This shift was hypothesised to be too slow for any conscious intention among the house builders, but rather a result of so-called pseudoneglect, or the tendency of a slight leftwards bias in the perception of parallel lines, mostly among the right-handed. @muller-scheessel2020 observed this gradual shift in the orientation of 17 houses in Vráble for which construction dates had been estimated based on Bayesian modelling of ^14^C dates. Their linear regression model for all these houses only gave a weak correlation however, and they argued for basing the model on the eight houses in the SW neighbourhood only, where the sampling strategy had been the most systematic. This gave a regression model defining house orientation as $0.129x - 651.016$ with correlation coefficient $r = 0.84$, $x$ being the modelled construction year BCE, which when applied inversely (solving for $x$) gives modelled construction year as $\frac{orientation + 651.016}{0.129}$. Use-life of houses in Vráble was modelled by @meadows2019 to a median of 27.5 years, a number which according to the same authors has varied greatly throughout the long history of Linear Pottery research. Extrapolating the model for construction year based on house orientation and adding this median duration onto all known houses in Vráble gives a settlement occupation span from 5297 to 4975 cal. BCE, which is in good agreement with the currently available and modelled ^14^C dates [though admittedly a few decades earlier, see @meadows2019]. For the distribution fitting analysis in this section I defined a set of 16 sample dates spaced 20 years apart from 5290 to 4990 within this timespan, and assigned houses to each sample accordingly. Since the duration between time samples is a few years shorter than the duration of houses, some houses were assigned to more than one sample. Furthermore, as with the analyses above, samples with 10 or less houses in them were filtered out, mainly since the distribution method then becomes too unreliable (this is already a quite generous allowance, Chapter \@ref(methods-distfit)), but also since it makes little sense to speak of hierarchy in a group of less than 10.

It must be noted that there are several caveats with this procedure of selecting out houses to be considered coeval. Firstly, the dating proxy method proposed by @muller-scheessel2020 is recent and not yet well established. It is based on linear regression of very few data points (8 houses in only one part of the settlement), all of which are not unique certain values but dates that are themselves also modelled. Though the publication does not specify it explicitly, it seems their model was fitted by ordinary least squares (OLS) regression, while systematic measurement uncertainty in the independent variable (the modelled ^14^C dates) should warrant more robust methods like orthogonal regression or probabilistic Monte Carlo methods. My use of a single value for house duration rather than the probability distribution that it really is, is also a simplification that could affect the results of the analysis. In general, longer house durations lead to more of the houses being coeval throughout the timespan of a settlement, while inversely shorter house durations lead to fewer coeval houses. There could also potentially be systematic relationships between house size and duration, e.g. that larger houses were occupied by more temporally stable households for several generations, putting in greater efforts to maintenance. These questions are not further pursued here however, and the results of this analysis must be considered preliminary.

## The Cucuteni-Trypillia culture complex {#trypillia}

The Cucuteni-Trypillia culture complex (CTCC) is also a large and well-defined archaeological culture, geographically somewhat less extensive than the Linear Pottery, but found over a longer chronological range of almost 2000 years from the early 5th millennium to around 3000 BCE, a period referred to as the Eneolithic or Chalcolithic/Copper Age within its regional context [@harper2021; @harper2023]. Covering a large area within modern Ukraine, Moldova and Romania from the foothills of the Carpathians in the west to the middle Dnieper in the east -- though inland from the coast of the Black Sea -- it was independently identified in the late 19th centrury at the eponymous sites of Trypillia on the right bank of the Dieper (Kyiv Oblast in Ukraine, then within the Russian Empire) and in Romania at Cucuteni (Iaşi County near the border to Moldova). Over much of the 20th century it has been debated whether they should be considered as two separate cultures, and today they are generally treated as one, or at least as closely related, as seen through their highly similar pottery traditions [@shatilo2021, p. 13 ff.]. One significant trait that does distinguish Cucuteni and Trypillia settlements, is a tendency of the former to form irregular grid layouts, while the latter form radial layouts of concentric circles. The CTCC settlements analysed in this study belong to the Trypillia group, and all show this characteristic settlement layout (further descriptions below). Trypillia sites are today most well known both among archaeologists and the wider public for their so-called "mega-sites" reaching up to about 3000 houses (Maidanetske) and covering areas up to 340 ha (Taljanki), settlement sizes that are otherwise unheard of within the European Neolithic, and have spawned a vivid debate as to whether they should be considered the earliest cities (more discussed below). Note that the spelling of Trypillia is transliterated to English from the Ukrainian toponyme, while it often -- especially in older literature -- is transliterated from the Russian equivalent Tripolje.

The near absence of stratified settlements within the Trypillia culture and the low availability of radiocarbon dating for Soviet and post-Soviet archaeologists, has lead to the elaboration of a highly complex sequence of relative chronology based on pottery typology. The general periodisation of Trypillia pottery and material culture -- with periods A, B, C and sub-divisions -- was originally made by Tatyana Passek in the 1930s and 40s, and still forms the backbone of the culture's chronology [Passek published only in Russian, but detailed English summaries with references are now available, e.g. @shatilo2021, pp. 18-26; @ryzhov2012]. Besides the temporal sequence, Trypillia material culture has subsequently been further divided into spatial sub-groups, notably to the main groups of the so-called Western and Eastern Trypillia cultures. The former (abbreviated to WTC in specialised literature) being primarily characterised by the prevalence of polychrome painted pottery (similar to Cucuteni pottery) while the Eastern group (ETC) is characterised by pottery with primarily incised and fluted decorations, building on traditions from the initial Precucuteni/Trypillia A phase. There is also a general long-term trend of migration and pioneer settlement from the south-west towards the north-east, and the ETC -- representing the earliest settled farmers in central Ukraine -- first appear after the initial stages of WTC and Cucuteni development [@harper2023, pp. 646-7]. Combining temporal and spatial dynamics of pottery styles, a number of geographically more-or-less stable local traditions have been proposed, termed "local groups" or "lines of development" for both the WTC and the ETC [@ryzhov2012; @diachenko2012; @diachenko2012a]. These are generally thought to represent somewhat coherent population groups, changing the location of their settlements at a relatively rapid rate, often in small steps but from time to time in larger leaps, leading to considerable geographical overlaps and mixing between different traditions (e.g. the WTC settlements presented below are actually further *east* than most of the ETC settlements).

It is clear that for much of its research history and until today, this richly detailed partitioning of the Trypillia into small local groups has gone hand-in-hand with a theoretical framework which has been largely rejected by anglophone archaeologists for a long time, namely the equation of pottery styles with ethnic or tribal groups [@shatilo2021, pp. 44-9]. While this academic tradition is still strong in Ukraine, as in much of south-east Europe more generally, it is increasingly being nuanced with more critical archaeological theory, e.g. by taking into account the numerous transition zones between groups of pottery styles, and acknowledging that other sources of material culture do not follow the pottery periodisation in a systematic and homogeneous way. Also, the substantial efforts made over the past decade or so of combining the existing Trypillia typo-chronology with modern AMS radiocarbon dates is currently leading to a more precise mapping of the spatio-temporal dynamics of specific pottery traits [@harper2023; @harper2021; @rassamakin2012].

Trypillia settlements in Ukraine and Moldova are to a large extent located within the forest-steppe eco-region -- a transition zone between the temperate forests further north and west, and the dry Pontic-Caspian steppe to the south and east. The forest steppe has a humid to semi-arid continental climate, and sustains naturally an open mixed oak forest. Recent archaeo-botanical and pedological studies have shown a temporal coincidence -- at least in the mega-site region -- between Trypillia occupation and the transition from more dense forest to open forest steppe, and @kirleis2016 suggest the initial formation of this "cultural steppe" with its associated Chernozem soil might result from deforestation induced by Trypillia settlers, which could also explain some of the frequently observed instability in settlement location over the period.

Despite the very long period of development of the Trypillia culture, subsistence economy is thought to have remained relatively stable. Animal husbandry was based mainly on cattle, pig and sheep/goat livestock, with some variation in proportions throughout the phases [@kirleis2016a]. Domestic species were supplemented with hunted game, with species depending on the local environment -- red deer and boar in woodland dominated areas, and occasionally horse nearer the steppe region. There is no evidence of horse domestication from Trypillia contexts. Hunting was most prevalent in the early phase (Trypillia A, up to almost 50% of faunal remains in some cases), and decreased gradually over time as husbandry became more prevalent. Plant cultivation included varieties of wheat and barley -- with the introduction of millet from Trypillia C -- as well as pulses (pea and lentil, bitter vetch in Trypillia A). Wild plants like hazelnut and various fruits were also consumed, some of which -- like apricot -- were of exotic origin and might have been cultivated.

It should also be noted that both of the eponymous sites are located toward the extremities of the area defined by CTCC sites, and though there are certain geographical biases in the documented material, it seems clear that the core area of settlement and population density for much of the culture's duration was in and near the river valleys of the Dniester and Southern Bug (mainly in present-day Moldova and south-western Ukraine). The Trypillia settlements included in this study are attributed to the B-II and C-I phases of the WTC and are located within a small micro-region in the Southern Bug-Dnieper interfluve, around the limit between the Cherkassy and Kirovograd oblasts, some 30-40 km east of Uman (Figure \@ref(fig:03-trypillia)). Based mainly on pottery style, one settlement -- Nebelivka -- is associated with the eponymous Nebelevskaya local group (phase B-II), while the three other settlements are associated with the related but slightly later Tomashovka group [phase C-I. See @shatilo2021, pp. 41-4 for a condensed overview of WTC local groups; @ryzhov2012; @diachenko2012a]. This small area has been the focal point of intense research over the last 15 years, and has -- among many other remarkable results -- led to the publication of complete or near-complete settlement plans, identified through geomagnetic imagery with unprecedented quality [@rassmann2014; @rassmann2016].

(ref:03-trypillia) Location of the Trypillia settlements analysed in this thesis, within the drainage basin of the Syniukha river, a small left tributary of the Southern Bug in central Ukraine. Terrain model data in public domain, courtesy of the U.S Geological Survey (<https://earthexplorer.usgs.gov>)

```{r 03-trypillia, fig.cap="(ref:03-trypillia)", fig.scap="Map with location of Trypillia sites"}
include_graphics("Results/fig03_trypillia.pdf")
```

### Sampled Trypillia settlements: The Syniukha basin

The four Trypillia settlements that were sampled and analysed in this thesis had been known by Ukrainian and Sovjet archaeologists for some time, having been subject to documentation through aerial photography and small-scale excavations mainly in the second half of the last century [see @ohlrau2020, pp. 36-60, 241-4; @kruts2012], but their settlement plans were first extensively documented through geomagnetic imagery from 2009 and onwards, as part of two collaborative research projects, one German-Ukrainian [Maidanetske, Moshuriv and Talne 3, @ohlrau2020, pp. 61-5; @rassmann2014] and one British-Ukrainian [Nebelivka, @hale2020; @chapman2014a]. This sample is admittedly very small, and other settlements from the same period have been documented in the area (Taljanki, Apolianka, Dobrovody and others), but with larger gaps in their settlement plans due to lacking access for the surveyors or less optimal site preservation. These four settlements are furthermore selected for their extreme variation in size -- from 3.1 ha at Talne 3 to 235 ha at Nebelivka -- despite being dated within a relatively short time period. The Trypillia B-II/C-I phases are currently broadly estimated to the range 4100-3500 BCE, and more precisely to 3950-3750 for the Nebelevskaya local group (Nebelivka) and 3850-3600 for the Tomashovskaya local group [Maidanetske, Moshuriv and Talne 3, @harper2021].

All four settlements are spatially organised according to the circular-radial layout characteristic to Trypillia settlements (see settlement plan fig.#). The two very large settlements -- Nebelivka and Maidanetske -- show some additional traits that have come to be known as characteristic for the so-called "mega-sites" [@hale2020; @ohlrau2020], namely:

-   Two or more concentric house circles separated by some space, creating a circular "main street"

-   A large open space in the centre of the settlement, entirely void of constructed features

-   A rectangular main plaza containing an exceptionally large building toward the east end of the settlement, inside the inner house circle (only a small portion of this building is preserved at Maidanetske, but its original presence is assumed by analogy to the other mega-sites)

-   Secondary large buildings located at highly visible spots at regular intervals within the main street (these, as well as the largest building, are generally termed "mega-structures" by the German team and "assembly houses" by the British team, see below)

-   Regular domestic houses constructed both on the outside and the inside of the main street, to varying degrees forming secondary streets (especially in Nebelivka) and plazas (Maidanetske)

-   Varying degrees of clustering of houses into quarters and neighbourhoods (Nebelivka terminology), presumably also reflected by the spatial distribution of assembly houses/mega-structures

-   Exceptionally large settlement size [@ohlrau2020, p. 27 proposes a threshold value of ca. 30 ha for all Trypillia phases]

-   Location within the forest-steppe, at the border to the steppe

The smallest settlement analysed here, Talne 3, consists of only a single circle, while the Moshuriv settlement has two. Both of them also have a single row of houses in the centre, contrarily to the mega-sites, wile Moshuriv also has its largest house placed in the eastern part of its inner house circle, similarly to the mega-sites. There is still much uncertainty regarding the internal temporal development of Trypillia settlements, as well as to their total durations (more discussion on this below). Based on the layout of smaller settlements like Talne 3 and Moshuriv -- seemingly growing outwards like year-rings in a tree -- it is hypothesised that this also was the case for mega-sites, i.e. that the inner circle would be established first, and the rest of the settlement plan emerging over some time [@ohlrau2022; @muller2016c]. In any case, it must be noted that mega-sites constitute an exception rather than the rule for Trypillia settlements, as they are only found within a very small region and over a relatively short range of the whole CTCC.

The further brief overview of house construction, architecture and social organisation is focussed mainly on the settlements within the Syniukha basin, though many of the traits are widely shared within larger parts of the Trypillia area.

### Trypillia house construction, architecture and social organisation

As with Linear Pottery houses, Trypillia domestic architecture has been amply documented through more than a century of numerous excavations. However, there are also three additional factors that have contributed to an even more detailed understanding of these latter houses:

-   The very common practice in Trypillia settlements of burning houses entirely at the end of their use life

-   Better post-depositional preservation than most comparable West-European soils, due to less intensive use of modern agro-industrial machinery, leaving architectural remains including living floors most of the time *in situ*, and

-   Numerous finds of detailed miniature clay models of Trypillia houses

Combined, these favourable conditions allow for a relatively precise archaeological reconstruction of how people built their houses and lived in them within the Trypillia culture [see @chernovol2012 for a detailed overview in English; @korvin-piotrovskiy2012]. On the other hand, Trypillia burials are very rare, and none have been found within the Syniukha basin, which has left large blind spots in the documentation regarding gender, kinship and family structures, in stark contrast with the situation in Linear Pottery research. It is generally assumed that one domestic house was used primarily by a single family unit, because of its layout and interior organisation (see below).

The main feature of house remains -- which is usually very visible on geomagnetic images -- is a compact clay floor of about 5 to 12 cm thickness interpreted as the main living space of the inhabitants. Imprints of wooden beams are regularly found on the downward side of floor fragments, and it is widely accepted that these floors were constructed on a raised platform of some height, leaving an open space below possibly for storage or even housing for livestock. This configuration is also regularly seen on clay house models, which also hint at the use of arched ceilings and gables. Walls were made mainly of wattle-and-daub, also seen in wattle imprints on fired daub fragments, but there is also occasional evidence of walls made from horizontal wooden beams -- possibly this technique was applied for wall foundations [@chernovol2012, p. 183]. The main living floor on the raised platform was presumably reached by a ladder from the outside, since openings are not found on the platform floors.

The interior organisation of rooms and installations was very homogeneous across the Trypillia complex, to the degree that Chernovol speculates that there were taboos associated with deviating from the strict norm [-@chernovol2012, p. 200]. For settlements in the Tomashovka local group (see above), the vast majority of houses were rectangular and about 3 to 4 times longer than wide, with the upper level composed of an entrance porch on one end, leading through a doorway to the main room. There, a clay oven was located immediately to the right of the entrance, while a trough and long podium followed the left side wall. By the short end opposite the entrance, there was a special installation usually interpreted as a domestic altar -- a clay disc about 1.6--2 m in diameter and 10--15 cm thick, often decorated with red and yellow paint and linear incised patterns. Tableware is regularly found at different parts of the house during excavation, notably near the altar and entrance, and the question of whether items were placed there in a meaningful way prior to house burning has been a matter of much discussion [@chernovol2012, p. 198]. Some more rare variations on this rather strict norm of house organisation include secondary altars either in the main room or the porch, as well as an additional annex -- possibly for craft activities -- appended to the front porch.

Loom weights are generally found within regular houses, most often on the raised platform near the altar, suggesting that weaving was a domestic activity. This is not the case for remains associated with flint knapping, which must have taken place at least outside regular domestic areas, and possibly within specialised workshops, some cases of which have been suggested within CTCC settlements [@chernovol2012, pp. 198-200]. The unusually fine (for the European Neolithic) Trypillia pottery seems to largely have been produced at sophisticated double-chamber kilns which are found across the mega-sites, indicating that this activity was either organised at a communal (quarter/neighbourhood) level or performed by specialised artisans -- or some combination of the two [@korvin-piotrovskiy2016; @costin1991]. Interestingly, @korvin-piotrovskiy2016 showed that these kilns were evenly distributed at sites like Taljanki and Petreni, while at Maidanetske they were more clustered and not corresponding to the distribution of mega-structures. Metal finds are very rare in Trypillia contexts, and it remains unclear to which extent metallurgy was actively practised or whether objects were imported from neighbouring regions in the lower Danube or the Carpathian basin. In any case, objects such as copper chisels/axes were probably not the commonly used tools for woodworking (polished stone adzes are more widely documented), at least not before the latest stages of the Trypillia culture [@gaydarska2020c, pp. 454-5]. A single small gold spiral, interpreted as a hair ornament, was found within the largest structure at Nebelivka, and is apparently the only gold find recorded in any Trypillia context [@gaydarska2020d].

Similarly to many South-East European Neolithic/Chalcolithic settlement contexts, Trypillia house remains are often -- or even usually -- heavily burnt at the end of their use-life. While earlier interpretations tended towards catastrophic site-ranging accidents or warfare (invasion from the steppe populations), more careful excavation and interpretation of house remains, combined with some experimental programs, have led to a consensus that the houses were intentionally set on fire individually as they fell out of use [@korvin-piotrovskiy2012; @johnston2020]. In the recent high-resolution geomagnetic images of Trypillia mega-sites, house remains were recorded as either "burnt" or "unburnt"/"eroded", depending on the intensity of the signal and thus how visible the remains were [@rassmann2016; @rassmann2014]. For the mega-sites the unburnt category was assigned to 21% of the houses at Maidanetske [610 of 2932 houses, @ohlrau2020, pp. 61-4] and 26% at Nebelivka [368 of 1445, @hale2020, pp. 124-9]. More recently this categorisation has been questioned, since excavated building remains labelled as unburnt also showed clear signs of firing at high temperatures. @pickartz2019 suggested that the so-called unburnt houses were simply constructed differently, using less clay and daub, and thus leaving less material after burning that would easily be identified through geomagnetics. Whether there was any functional difference between these two house categories remains an open question.

The special category of much larger houses, generally referred to as "mega-structures" and primarily known from the "mega-sites", was first discovered in 2009 through geomagnetic imagery, and has received much attention since [comparative overview in @hofmann2019; see also @nebbia2018; @chapman2016; @rassmann2016; @burdo2016; @muller2016a; @chapman2014; @rassmann2014]. These are recognised by a number of architectural features that clearly distinguish them from the regular domestic houses described above [@hofmann2019, p. 35 ff.]. Firstly, they usually have twice the floor area or more compared to regular houses -- buildings referred to as "mega-structures" range in sizes from about 100 m^2^ up to 1.200 m^2^ for the largest known examples (in Dobrovody and Nebelivka) -- and are spatially distributed in characteristic ways within the settlements. They are not constructed on raised platforms, and in many cases they are only partially if at all roofed (e.g. with a large walled front yard or atrium). Only a few of these buildings have yet been excavated -- @hofmann2019 enlist four: one each in Maidanetske, Nebelivka and Dobrovody (all representing Trypillia B-II/C-I mega-sites in the Syniukha basin, central Ukraine) as well as one in Baia (Precucuteni, north-east Romania). The excavated mega-structures in Dobrovody and Nebelivka both represented the largest structures within their settlement, and while the former was interpreted as an entirely unroofed rectangular enclosure, several key aspects of the internal architecture of the latter -- the existence of an upper floor, supporting vertical posts and galleries, number of use phases, burning events -- have been subject to highly conflicting views [@gaydarska2020b; @chapman2016; @burdo2016; @chapman2014]. The excavated mega-structure at Maidanetske was not the largest at the site, but both its size and positioning clearly distinguished it from domestic houses. It was interpreted by the excavators as being sectioned in two almost equally sized parts, one roofed and one walled open yard [@hofmann2019, pp. 13-29].

All of the three excavated mega-structures in the Syniukha basin yielded rather regular domestic purpose finds, but in much lower quantities than what would be expected from their sizes if they were permanently inhabited. This, along with the apparent absence of installed ovens, has led the involved research teams to interpret them as being non-domestic in function, i.e. that they did not serve as the homes of ruling-class elites or chiefs. However, what exactly *did* take place there remains a more open question. The Nebelivka mega-structure housed as much as seven cruciform clay platforms analogous to the "altars" of domestic houses, but some of them being much larger. The Ukrainian part of the excavation team at Nebelivka has advocated an interpretation of a two-storey building, with the inner room of the upper floor resembling a dramatically up-scaled version of a domestic house, suggesting that this part of the structure could be well suited for permanent occupation [@burdo2016, p. 111], while the mentioned altar platforms -- showing clear traces of fire-related use -- would be located on the lower floor. The authors do not hesitate to characterise the structure as a whole as a communal temple. The British part of the Nebelivka excavation team has favoured an interpretation where only the middle part of the structure would be roofed, and where no rooms would be suited for year-round occupation [@chapman2016]. While they are more prudent regarding the interpretation of the raised clay platforms as altars (they prefer the more neutral term "Raised Area"), they in turn do not hesitate to designate the complex as an "Assembly House", an interpretive term first suggested for the Trypillia mega-structures by Cahokia specialist Timothy Pauketat [@hale2020, p. 123].

Somewhat curiously, while the two parts of the excavation team largely disagree in their architectural and social interpretations of this exceptional building, they do seem to agree on the one hand that it was not the residence of an elite leadership, while on the other hand it holds several characteristics of such, including its monumentality in size, elevation and visibility, restricted access, the presence of large storage vessels and stamp seals ("tokens"), as well as special finds and long-distance imports not found elsewhere on site (or hardly within the entire CTCC), like a small gold spiral and graphite-decorated pottery [@gaydarska2019, pp.107-9; @gaydarska2020d]. One could almost ask if the same structure found in a later context, like the Bronze Age Aegean, would not be immediately recognised as the residence of a *wanax*. The finds are certainly somewhat scarce in quantity, but considering that the ritual (non-accidental) destruction of the building at the end of its use-life is beyond reasonable doubt, this can easily be explained either by a prior removal of administratively important inventory as well as valuable items, leaving only the objects required for the burning ceremony, either by the building's inhabitants themselves or pillaged by others if the elites were removed from office. If the structure was really used communally for seasonal feasts and ceremonies, one would normally expect far more broken tableware and feasting deposits (insert Hayden ref here#). The main recurring argument against a top-down rule at Nebelivka and other Trypillia mega-sites seems to be the lack of evidence from monumental graves [e.g. @nebbia2018]. However, the evidence of burial practices is almost entirely lacking from the Trypillia as a whole, and in particular within the mega-site region, so it seems clear that this is a part of their cultural practice that is currently far from well understood. Though I do admittedly not have first-hand field experience with Trypillia settlements, I am inclined to hold that the burden of proof in this case should be on those who claim that Trypillia mega-sites reflect an "egalitarian society" [@nebbia2018, p. 51; @gaydarska2020c; @graeber2021, pp. 288-97].

The smaller secondary mega-structure (placed within the "main street" of the settlement) excavated by the German-Ukrainian project at Maidanetske did not yield special exotic finds, but was similarly characterised by a striking paucity of more standard finds categories [@hofmann2019, pp. 13-29]. Similarly to the large Nebelivka mega-structure, this structure was also seemingly partitioned in a roofed and an open section, but with entrances from the outside to both parts. The activities taking place within the structure, judging from the finds inventory, were more of household-economic rather than ritual or administrative character, like weaving, cereal processing and storage, but also food preparation and meat consumption connected to an open fireplace. The absence of ovens and the generally lighter architecture compared to domestic houses were seen as further indications of economic rather than domestic function, and arguing from numerous ethnographic analogies the research team proposed that these large main street buildings were polyvalent community structures for collective use. Their regular placement within the mega-sites serves as a strong argument for a settlement organisation of socio-economically separate quarter communities, integrated together at the site-level in some way or another through the largest building and its main plaza. @hofmann2019 furthermore noted that the potential use-groups of each secondary mega-structure (number of households) by any measure would largely exceed the numbers seen in ethnographic analogies, arguing that there could well be also a third level of social integration reflected through house clusters (the single household being a fourth level). However, the numbers of coeval houses at Maidanetske and other mega-sites are yet far from being securely established.

### Are the maga-sites cities?

As a last note on Trypillia social organisation, mention must be made of the ongoing debate about whether the mega-sites should be considered as cities rather than large villages. If so -- and given that they pre-date the earliest cities in Mesopotamia by several centuries -- they should be recognised as the earliest cities in the world. Definitions of cities or urban settlements are numerous but generally follow that of @wirth1938, who listed their characteristic criteria as being large, dense and permanently settled by heterogeneous individuals -- i.e. exhibiting some socio-economic variability and thus exchange of services and resources between the inhabitants, as well as a functional complementarity with surrounding rural areas. For the Trypillia mega-sites, the debate has thus been largely focussed on population estimates, craft specialisation, economic diversity and hierarchical administration, as well as their functional relations with the hinterland. Population estimates have varied widely, from some hundreds to tens of thousands living coevally at the sites [see e.g. @gaydarska2020c, p. 433 for an overview of estimates for Maidanetske]. Surprisingly, the researchers who have advocated for the lowest coeval populations [the "minimalist view" in the terms of @gaydarska2020e] are also those who have insisted the most on labelling the sites as urban. Arguing from their evidence of an apparently limited human impact on the surrounding environment (e.g. degrees of erosion, deforestation, pollen records) over the duration of Nebelivka, @gaydarska2020c have proposed that the site would likely not be permanently occupied [see also @gaydarska2020e; @gaydarska2019; @nebbia2018]. They developed three main alternatives for explaining this paradox of a high population and low environmental impact [summarised in @gaydarska2020e, p. 35; @gaydarska2019, pp. 111-3]:

-   The Assembly model (colloquially referred to among the excavators as the "Burning Man" model), of a yearly gathering of disparate communities for a short duration, to partake in ceremonial exchanges of information, goods, partners etc.
-   The Pilgrimage model, where the location of the settlement would have some great religious significance to the population of a much wider region
-   The Distributed Governance model, where each quarter community would take turns on a yearly basis of the responsibility for acquiring necessary resources for the others from a wider economic region

According to the two first models, only a small "caretaker" population would be permanently settled at the mega-site. The third model would posit some middle-ground allowing for a somewhat larger coeval population, but with the organisation of substantial resource imports from elsewhere. All three models have some support from ethnographic analogies [amply discussed in @graeber2021], but have also been criticised for being fanciful narratives lacking grounding in archaeological evidence [e.g. @ohlrau2020, pp. 274-6; @ohlrau2022]. The rather agrarian character of finds inventories from Nebelivka and other mega-sites -- typical of what is usually labelled as Neolithic or Chalcolithic societies -- as well as the low settlement density following from the models of only partial occupation, led the researchers to interpret the Trypillia mega-sites as examples of *low-density* *urbanism*, a term promoted by @fletcher2011 to describe agrarian dispersed cities, primarily within the tropics, like Ankor and Tikal [@chapman2016a].

The German-led investigations at the mega-sites have on the one hand proposed higher coeval population estimates, higher densities and generally refuted the seasonality hypotheses suggested by the British team -- also questioning the environmental evidence serving as its main argument, e.g. @ohlrau2022, p. 85 -- and on the other hand been more reluctant to characterise them as urban [@muller2016a; @muller2016b; @muller2016c]. To @muller2016b, typically urban traits that are seen at late 4^th^ millennium Uruk in Mesopotamia, like the spatial separation between a religious-administrative and domestic districts as well as clear signs of division of labour, are not present at the Trypillia mega-sites. The perhaps most "maximalist" views have been held by Ukranian specialists who, grounded in the typo-chronological tradition of following the movements of population groups at the regional scale, with added results from modelled population dynamics, have pointed to the large scale (largely coeval) but short-livedness and apparent instability of the mega-sites, holding that the very large groups of people could not be sustained by the environment for a long time at the same location, while unoccupied land suited for resettlement was readily available [@shatilo2021; @diachenko2017; @diachenko2015; @diachenko2012; @diachenko2012a].

While these studies have tackled the uniqueness of the mega-site phenomenon differently, inn all cases it seems clear that for a number of reasons these settlements do not fit classic definitions of cities in a straight-forward way. More recently, @sindbæk2022 proposed the term *anomalocivitas* to characterise various types of settlements known through history and prehistory for which the urban definition is seen as problematic. Acknowledging that efforts to shoehorn archaeological material into inadequate definitions are often based on teleological fallacies, preventing researchers from seeing the settlements for what they really are, he constructed a framework of definitions allowing for more variability and connecting the "lack" of certain urban features with specific socio-economic functions [@sindbæk2022, p. 21-2]. In this sense, Trypillia mega-sites are similar to Iron Age *oppida* like Bibracte and Manching, or Viking trade *emporia* like Ribe and Kaupang, in that they all have low values of one variable central to urbanism -- low size in the case of *emporia*, low density in the case of *oppida*, and low heterogeneity in the case of the mega-sites. The low socio-economic heterogeneity is interpreted [e.g. @ohlrau2022; @muller2016b] as a result of the agglomeration of essentially similar village populations from within a small region -- populations who had not already developed a high degree of economic specialisation unlike in the Near East Chalcolithic. @hofmann2019 see a gradual tendency towards centralisation and monopolisation of power within the Trypillia mega-sites over the period of their duration (secondary mega-structures becoming fewer and smaller, the spatial distribution of pottery kiln becoming increasingly clustered), and argue that this, rather than ecological degradation, caused their ultimate disintegration in the Trypillia C-i/C-II transition. As for now, the most unresolved question regarding the Trypillia mega-site anomaly, seems to be why they appeared and grew so large in the first place.

## Reading site plans from geomagnetic imagery

The unprecedented quality of the recently published geomagnetic plans of Linear Pottery and Trypillia settlements [@rassmann2016; @rassmann2014; @chapman2014a; @muller-scheessel2020], as well as the theoretical and culture-historical considerations presented at the beginning of this chapter, make these two study areas ideal as cases for testing the performance of the fractal analysis methodology for quantifying levels of social hierarchy. For this thesis, spatial data in the form of shape-files were generously shared by the GIS specialists of the different research projects: Niels Müller-Scheeßel [Kiel university, @muller-scheessel2020] for the Linear Pottery sites in the Žitava valley, René Ohlrau [Kiel university, @ohlrau2020] for Maidanetske, Moshuriv and Talne 3, and Duncan Hale [Durham university, @hale2020] for Nebelivka. The house outlines were thus identical to those presented in the cited publications. However, this also means that they were drawn by (at least) three different people, which may have lead to certain biases between the data sets. Furthermore, Linear Pottery and Trypillia houses are not identified on geomagnetic images in the same way: the former are seen as "empty" spaces between the longitudinal pits which are generally well visible [@winkelmann2020], while the latter are seen more directly by the burnt clay platform forming the floor area [visible to varying degrees, as already discussed above, @pickartz2021; @pickartz2019]. Both culture groups are known to allow for very little overlap between houses -- an issue which would be far more problematic e.g. with Balkan or Anatolian Neolithic villages. The data to be extracted from these files were house sizes for Chapter \@ref(results-distfit) and the house outlines themselves as images for Chapter \@ref(images-results). Both the extraction of house (polygon) sizes and settlement layout raster images were done using standard functions in QGIS 3.20.1 (see Section \@ref(images-procedure) for more details on image preparation).

## Ethnographic and synthetic data

A challenge with applying new methods onto any material with the goal of using the obtained results as proxies for some unknown or little known variable (in this case the level of social hierarchy), is that this use of the method must in some way be validated by also applying it on material where the variable is known. At the onset of this PhD project my intention was to compare results obtained from the archaeological samples with results from ethnographic samples. Using the settlement plans of villages and settlements where the actual degree of social hierarchy had been analysed and documented by ethnographers would have been a valuable contribution to this study. However, it quickly became clear that such settlement plans, with the necessary degree of detail and accuracy, are hard to come by. For comparability with Neolithic societies, the ethnographic sample should ideally consist of settlements having been minimally transformed by modern industrialisation -- archaeologists specialising in Neolithic Europe tend to prefer comparisons with Sub-Saharan African or Austronesian village societies, preferably documented before the second half of the 20^th^ century. Unfortunately, while such documentation is abundant, in the vast majority of cases settlement layouts and house sizes are either entirely missing, or documented in a highly approximate manner, inhibiting any meaningful comparison with the sampled archaeological settlement plans. Ethnographers have simply not been very interested in mapping villages by precise measurement in the past. However, although not directly transferable to the quantitative analyses proposed in this thesis, relevant ethnographic literature is to some extent reviewed in Sections \@ref(house-sizes-ethno) and \@ref(layout-ethno).

An alternative approach was therefore adopted, using synthetically generated data (i.e. computer-simulated), both for size distributions and spatial layouts (Chapters \@ref(methods-distfit) and \@ref(images-methods)). The advantage of this approach to the ethnographic one, is that it allows for setting parameter values entirely freely and thus testing the methods over wide and regular ranges of data. The disadvantage is that the connection to actual real-world social systems remains somewhat uncertain, as the synthetic data simulates the material culture (the settlement plans) and not the social system that generated it. This gap between the data and interpretation can to some extent be filled by considering theoretical perspectives. While the theory of size distributions and their generative social mechanisms is well developed (Section \@ref(distributions)), this is much less the case for the theory regarding spatial layouts and textures of social organisation (Section \@ref(layouts)).
