# Material and data: social complexity in the European Neolithic {#material}

## Studying social hierarchy in archaeology and prehistory. {#complexity-archaeo}

Grave goods

Example of Varna: Lichardus (1991; also Gimbutas), as discussed in @kadrow2013, proposed that the social inequality observed at Varna in the mid-5th millennium, was already then adopted from pastoral North Pontic steppe cultures (*check this and delete if unsure*); cf. @chapman2006 -- three social levels interpreted from grave goods, but not clear if these are really discrete or constructed on a continuum. The term *fractal* is here used purely as a metaphor for complex personhood structures, rather in opposition to the described social hierarchy. Biehl and Marciniak in Diehl 2000: no hierarchy in Varna. Response by Hayden 200?

Burial monuments

Architecture @rathje1982 -- social stratification (qualitatively) through architecture and burials (and life expectancy). Short but to-the-point article.

The denominator problem

The use of ethnography

Other approaches (osteological, isotopes, craft specialisation **refs**)

This project: house-size distributions and settlement layouts (details in subsequent chapters), just very short argumentation.

Comparative approach: Neolithic technology (not bronze axes), wood and wattle-and-daub architecture, (near) complete settlement plans/extensive documentation: Trypillia and Linear Pottery

For this project I opted for the use of houses and built environments as proxy for social hierarchy. In this way I hope to largely avoid the denominator problem associated with burials. While it is true that for many archaeological culture groups in late prehistory habitats are poorly preserved and hard to discover, leaving us still with a limited understanding of them (the case in many Michelsberg, Corded Ware and Bell Beaker groups, only to mention a few **check**), in groups where habitats are well preserved, there is little reason to suspect that the available record would not cover the whole range of social statuses if these societies were hierarchical. Unlike burials, every individual in a sedentary society -- with few exceptions like homeless persons in more recent urban contexts -- will normally have at least one fixed place to stay overnight, and these homes will in most cases be constructed within the same fundamental framework of techniques and building materials, depending more on culture specific traditions and environmental factors than on social status (\#**citation?**). As an example, in a society where mudbrick is the main building material, like in the Neolithic Near East and Anatolia, nearly all constructions are made in mudbrick, regardless of the social status of the inhabitants. In continental Europe north of the Alps, wattle-and-daub construction was the almost exclusive building technique for any architectural feature from the early Neolithic until the Roman conquest, and well into the Middle Ages north of the *limes*. One can of course enumerate exceptions, but more importantly houses are in any stratified society also a marker of social status, which can be exhibited in a range of ways, from decorations, use of more precious raw materials as well as size. That is precisely the reason for using houses as a proxy for social status and hierarchy in archaeological settings. But the point here is that there should be little taphonomic differentiation between groups of high and low status within a given archaeological context, at least in prehistory, and at least not as much as can be expected for burials, meaning that we can expect to find samples of houses that are representative of the social structure of the archaeological culture in question. On sites where there is taphonomical loss of architectural structures, as long as the overall building tradition is homogeneous, there is no reason this loss should affect one segment of the society more or less than others.

Some caveats do remain, however, for the use of houses as proxy of social status. Firstly, there may be a documentation bias favouring larger houses, since they may be easier to discover both during excavation and in remote-sensing surveys (\#**citation?**). In samples from originally very skewed house-size distributions, there may also be a further taphonomic bias towards large houses, since smaller houses -- being far more numerous -- are statistically in greater risk of being affected by post-depositional disturbances. Both of these biases are hard to evaluate empirically, though computer modelling could potentially give indications of their importance. This, however, is not within the scope of the present study.

A second, and maybe more important issue, is that of the contemporaneity of houses. When the goal is to investigate the social structure of a settlement as reflected in its architecture -- be it through the size distribution of buildings or their spatial layout -- all the analysed features should ideally have been in use at the same point in time. This is however very hard to achieve in most archaeological settings, and many researchers choose to either ignore the issue, or to accept a temporal resolution that is far wider than what their research questions should logically allow for [@perreault2019]. One way of limiting this problem is to select study samples with little to no stratigraphic overlap, which might indicate short occupation span, though as shown below this indicator can also be deceiving. For both study areas selected for this thesis -- the Linear Pottery in south-west Slovakia and the B2/C1 Trypillia of central Ukraine -- settlement plans show very little overlap between houses, even though some of them probably developed over more than three centuries, as shown by radiocarbon dates and modelling (see below). Such settlement plans may be impossible to differentiate into separate coeval time samples without precise dating of construction and abandon of every individual house, counting in the thousands on the Trypillia mega-sites (an alternative method is presented only for the Linear Pottery settlements in Section \@ref(lbk)). On the other hand, the fact that there is so little overlap between houses despite temporal differences clearly illustrates how these settlement plans emerged over time, not as a *tabula rasa* in each generation but rather with new constructions respecting the location and orientation of older ones long after abandonment. Though such practices are indeed interesting, it is not at all obvious, however, to which extent they may reflect or even relate to social factors such as hierarchy. With settlement types with much higher degree of stratigraphic overlap, like the tell sites of the Balkan and Near Eastern Neolithic traditions, it may be easier to distinguish more or less coeval occupation phases, but they are again harder to document extensively -- because of the high density of archaeological finds and features, excavation surfaces typically cover only very small portions of the settlement, while remote sensing performs less well and not allowing for temporal disentanglement of constructed features (\#**citation**). In any case, the issue of temporal resolution of the data and its influence on analytical results is crucial, and will be discussed repeatedly throughout this thesis, with a summary in Chapter \@ref(disc-methods).

Comparative technology

Meaning of "house"

In the following I propose a very brief overview of the general archaeological contexts of the two cases analysed in this thesis, namely the Linear Pottery and the Trypillia cultures. Both have been extensively studied and enjoy long and rich research histories, and any attempt at a comprehensive overview here would be futile. In stead, I focus on the aspects that are the most relevant for the present study -- in particular of architecture, settlement layout and social organisation -- and address the interested readers to more complete reviews. In the final sections of this chapter I provide some discussion on the nature and limits of the data material used here, which consists mainly of geomagnetic imagery with expert interpretation of archaeological features, as well as synthetic data for simulations.

## The Linear Pottery culture complex {#lbk}

General intro to the culture: @bickle2013 (online!); @whittle2014.

Formation, characteristics -- striking difference from earlier Mesolithic cultures, hence the term "Neolithic Revolution" pinned by Vere Gordon Childe for the seemingly abrupt transition to agriculture in continental Europe with what he called the Danubian Ia **ref**. @whittle2022, check with Trigger!

expansion trajectory: east: @saile2020,

Questions of contact and even creolisation with local hunter-gatherer groups in the margins of the Linear Pottery settlement area -- Limbourg and La Hoguette pottery groups in northern France and Belgium, Swifterbandt in the Netherlands??, Ertebølle in northern Germany??, Bug-Dniester in Ukraine and Moldova [@saile2020] -- as well as with other already established Neolithic groups in central France [@roussot-larroque2022] and the lower Danube [@saile2020], remain hotly debated

End of LBK as collapse? The whole culture as an adaptive cycle: @gronenborn2014.

Organisation of Linear Pottery society: egalitarian or hierarchic? Cemeteries: @jeunesse2022, check @gaydarska2023 and @augereau2021. Linear Pottery architecture and house construction

origin of the longhouse: @bánffy2020, @coudart2015, @last2015

house typology, construction, and use: @modderman1970, @coudart1998

house orientation: "origin" hypothesis: @bradley2001

Number of coeval houses in LBK villages, evaluated to a mean of 5, sd. 3 [@coudart1998, p. 91]. Furthermore, based on ethnographic analogies she estimated a typical Linear Pottery village of 3 to 8 houses (of which 1 or 2 longer than 22 m), to include on average approximately 150-260 inhabitants, though fewer (up to around 80 people) for the smallest hamlets of 3 houses.

Patrilocality in LBK based on Sr isotopes: @hrncir2020a (read this more carefully), and other data (check) @schiesberg2016.

### Sampled Linear Pottery settlements: the Žitava valley

(ref:03-zitava) Location of the Linear Pottery settlements analysed in this thesis, located in the river valley of the Žitava, a small left tributary (via the Nitra and Váh rivers) of the Danube in south-west Slovakia. Terrain model data in public domain, courtesy of the U.S Geological Survey (<https://earthexplorer.usgs.gov>)

```{r 03-zitava, fig.cap="(ref:03-zitava)", fig.scap="Map with location of analysed Linear Pottery sites"}
include_graphics("Results/fig03_zitava.pdf")
```

fds

### Time samples of the Vráble settlement (moved from Chap. 6, review to check it fits)

Separating an archaeological settlement plan into samples with temporally coeval structures only is a demanding exercise, and more so the larger the settlement. The best and most reliable method is to date every single structure in the settlement, but this is usually impossible both in theory, since not all structures yield datable material, and in practice when considering the costs involved. The more realistic approach is to sample a smaller amount of structures throughout the settlement for dating and modelling, and then extrapolate the results to get a more or less rough overview of how much, proportionally, of the site was occupied at the same time. This has been done for the Trypillia mega-sites, indicating that the maxima of coeval habitation would be at approximately 33% of houses at Nebelivka [@millard2020, pp. 253-6; @muller2022] and 52% at Maidanetske [@ohlrau2020, pp. 233-5; @ohlrau2022, pp. 86-8]. Even though some general tendencies of the spatio-temporal development of these settlement plans have been proposed [@shatilo2021, p. 247; @muller2022, pp. 218-9], it is for now impossible to extract precise temporal samples of coeval house-size distributions here. This also goes for the image analysis presented in Chapter \@ref(images-results) [see also @bruvollforthcoming]. The smaller sites of Moshuriv and Talne 3 are only dated through relative chronology of surface finds and thus attributed as a whole to their chronological phase, though their small sizes and ordered layout would suggest that most houses could be coeval [@ohlrau2020, pp. 241-4].

For houses in Linear Pottery settlements, a special dating proxy was recently developed, based on the observation that within this cultural context houses were constructed to be parallel to pre-existing ones, but with a slow and gradual counter-clockwise shift in orientation of approximately 1.3 degrees per decade [@muller-scheessel2020]. This shift was hypothesised to be too slow for any conscious intention among the house builders, but rather a result of so-called pseudoneglect, or the tendency of a slight leftwards bias in the perception of parallel lines, mostly among the right-handed. @muller-scheessel2020 observed this gradual shift in the orientation of 17 houses in Vráble for which construction dates had been estimated based on Bayesian modelling of ^14^C dates. Their linear regression model for all these houses only gave a weak correlation however, and they argued for basing the model on the eight houses in the SW neighbourhood only, where the sampling strategy had been the most systematic. This gave a regression model defining house orientation as $0.129x - 651.016$ with correlation coefficient $r = 0.84$, $x$ being the modelled construction year BCE, which when applied inversely (solving for $x$) gives modelled construction year as $\frac{orientation + 651.016}{0.129}$. Use-life of houses in Vráble was modelled by @meadows2019 to a median of 27.5 years, a number which according to the same authors has varied greatly throughout the long history of Linear Pottery research. Extrapolating the model for construction year based on house orientation and adding this median duration onto all known houses in Vráble gives a settlement occupation span from 5297 to 4975 cal. BCE, which is in good agreement with the currently available and modelled ^14^C dates [though admittedly a few decades earlier, see @meadows2019]. For the distribution fitting analysis in this section I defined a set of 16 sample dates spaced 20 years apart from 5290 to 4990 within this timespan, and assigned houses to each sample accordingly. Since the duration between time samples is a few years shorter than the duration of houses, some houses were assigned to more than one sample. Furthermore, as with the analyses above, samples with 10 or less houses in them were filtered out, mainly since the distribution method then becomes too unreliable (this is already a quite generous allowance, Chapter \@ref(methods-distfit)), but also since it makes little sense to speak of hierarchy in a group of less than 10.

It must be noted that there are several caveats with this procedure of selecting out houses to be considered coeval. Firstly, the dating proxy method proposed by @muller-scheessel2020 is recent and not yet well established. It is based on linear regression of very few data points (8 houses in only one part of the settlement), all of which are not unique certain values but dates that are themselves also modelled. Though the publication does not specify it explicitly, it seems their model was fitted by ordinary least squares (OLS) regression, while systematic measurement uncertainty in the independent variable (the modelled ^14^C dates) should warrant more robust methods like orthogonal regression or probabilistic Monte Carlo methods. My use of a single value for house duration rather than the probability distribution that it really is, is also a simplification that could affect the results of the analysis. In general, longer house durations lead to more of the houses being coeval throughout the timespan of a settlement, while inversely shorter house durations lead to fewer coeval houses. There could also potentially be systematic relationships between house size and duration, e.g. that larger houses were occupied by more temporally stable households for several generations, putting in greater efforts to maintenance. These questions are not further pursued here however, and the results of this analysis must be considered preliminary.

## The Cucuteni-Trypillia culture complex {#trypillia}

The Cucuteni-Trypillia culture complex (CTCC) is also a large and well-defined archaeological culture, geographically somewhat less extensive than the Linear Pottery, but found over a longer chronological range of almost 2000 years from the early 5th millennium to around 3000 BCE, a period referred to as the Eneolithic or Chalcolithic/Copper Age within its regional context [@harper2021; @harper2023]. Covering a large area within modern Ukraine, Moldova and Romania from the foothills of the Carpathians in the west to the middle Dnieper in the east -- though inland from the coast of the Black Sea -- it was independently identified in the late 19th centrury at the eponymous sites of Trypillia on the right bank of the Dieper (Kyiv Oblast in Ukraine, then within the Russian Empire) and in Romania at Cucuteni (Iaşi County near the border to Moldova). Over much of the 20th century it has been debated whether they should be considered as two separate cultures, and today they are generally treated as one, or at least as closely related, as seen through their highly similar pottery traditions [@shatilo2021, p. 13 ff.]. One significant trait that does distinguish Cucuteni and Trypillia settlements, is a tendency of the former to form irregular grid layouts, while the latter form radial layouts of concentric circles. The CTCC settlements analysed in this study belong to the Trypillia group, and all show this characteristic settlement layout (further descriptions below). Trypillia sites are today most well known both among archaeologists and the wider public for their so-called "mega-sites" reaching up to about 3000 houses and covering areas up to #FILL ha, settlement sizes that are otherwise unheard of within the European Neolithic, and have spawned a vivid debate as to whether they should be considered the earliest cities (more discussed below). Note that the spelling of Trypillia is transliterated to English from the Ukrainian toponyme, while it often -- especially in older literature -- is transliterated from the Russian equivalent Tripolje.

The near absence of stratified settlements within the Trypillia culture and the low availability of radiocarbon dating for Soviet and post-Soviet archaeologists, has lead to the elaboration of a highly complex sequence of relative chronology based on pottery typology. The general periodisation of Trypillia pottery and material culture -- with periods A, B, C and sub-divisions -- was originally made by Tatyana Passek in the 1930s and 40s, and still forms the backbone of the culture's chronology [Passek published only in Russian, but detailed English summaries with references are now available, e.g. @shatilo2021, pp. 18-26; @ryzhov2012]. Besides the temporal sequence, Trypillia material culture has subsequently been further divided into spatial sub-groups, notably to the main groups of the so-called Western and Eastern Trypillia cultures. The former (abbreviated to WTC in specialised literature) being primarily characterised by the prevalence of polychrome painted pottery (similar to Cucuteni pottery) while the Eastern group (ETC) is characterised by pottery with primarily incised and fluted decorations, building on traditions from the initial Precucuteni/Trypillia A phase. There is also a general long-term trend of migration and pioneer settlement from the south-west towards the north-east, and the ETC -- representing the earliest settled farmers in central Ukraine -- first appear after the initial stages of WTC and Cucuteni development [@harper2023, pp. 646-7]. Combining temporal and spatial dynamics of pottery styles, a number of geographically more-or-less stable local traditions have been proposed, termed "local groups" or "lines of development" for both the WTC and the ETC [@ryzhov2012]. These are generally thought to represent somewhat coherent population groups, changing the location of their settlements at a relatively rapid rate, often in small steps but from time to time in larger leaps, leading to considerable geographical overlaps and mixing between different traditions (e.g. the WTC settlements presented below are actually further *east* than most of the ETC settlements).

It is clear that for much of its research history and until today, this richly detailed partitioning of the Trypillia into small local groups has gone hand-in-hand with a theoretical framework which has been largely rejected by anglophone archaeologists for a long time, namely the equation of pottery styles with ethnic or tribal groups [@shatilo2021, pp. 44-9]. While this academic tradition is still strong in Ukraine, as in much of south-east Europe more generally, it is increasingly being nuanced with more critical archaeological theory, e.g. by taking into account the numerous transition zones between groups of pottery styles, and acknowledging that other sources of material culture do not follow the pottery periodisation in a systematic and homogeneous way. Also, the substantial efforts made over the past decade or so of combining the existing Trypillia typo-chronology with modern AMS radiocarbon dates is currently leading to a more precise mapping of the spatio-temporal dynamics of specific pottery traits [@harper2023; @harper2021; @rassamakin2012].

Trypillia settlements in Ukraine and Moldova are to a large extent located within the forest-steppe eco-region -- a transition zone between the temperate forests further north and west, and the dry Pontic-Caspian steppe to the south and east. The forest steppe has a humid to semi-arid continental climate, and sustains naturally an open mixed oak forest. Recent archaeo-botanical and pedological studies have shown a temporal coincidence -- at least in the mega-site region -- between Trypillia occupation and the transition from more dense forest to open forest steppe, and @kirleis2016 suggest the initial formation of this "cultural steppe" with its associated Chernozem soil might result from deforestation induced by Trypillia settlers, which could also explain some of the frequently observed instability in settlement location over the period.

Despite the very long period of development of the Trypillia culture, subsistence economy is thought to have remained relatively stable. Animal husbandry was based mainly on cattle, pig and sheep/goat livestock, with some variation in proportions throughout the phases [@kirleis2016a]. Domestic species were supplemented with hunted game, with species depending on the local environment -- red deer and boar in woodland dominated areas, and occasionally horse nearer the steppe region. There is no evidence of horse domestication from Trypillia contexts. Hunting was most prevalent in the early phase (Trypillia A, up to almost 50% of faunal remains in some cases), and decreased gradually over time as husbandry became more prevalent. Plant cultivation included varieties of wheat and barley -- with the introduction of millet from Trypillia C -- as well as pulses (pea and lentil, bitter vetch in Trypillia A). Wild plants like hazelnut and various fruits were also consumed, some of which -- like apricot -- were of exotic origin and might have been cultivated.

It should also be noted that both of the eponymous sites are located toward the extremities of the area defined by CTCC sites, and though there are certain geographical biases in the documented material, it seems clear that the core area of settlement and population density for much of the culture's duration was in and near the river valleys of the Dniester and Southern Bug (mainly in present-day Moldova and south-western Ukraine). The Trypillia settlements included in this study are attributed to the B-II and C-I phases of the WTC and are located within a small micro-region in the Southern Bug-Dnieper interfluve, around the limit between the Cherkassy and Kirovograd oblasts, some 30-40 km east of Uman (Figure \@ref(fig:03-trypillia)). Based mainly on pottery style, one settlement -- Nebelivka -- is associated with the eponymous Nebelivka local group (phase B-II), while the three other settlements are associated with the related but slightly later Tomashovka group [phase C-I. See @shatilo2021, pp. 41-4 for a condensed overview of WTC local groups; @ryzhov2012]. The further brief overview of house construction, architecture and social organisation is focussed mainly on the settlements within this area, though many of the traits are widely shared within larger parts of the Trypillia area.

(ref:03-trypillia) Location of the Trypillia settlements analysed in this thesis, within the drainage basin of the Syniukha river, a small left tributary of the Southern Bug in central Ukraine. Terrain model data in public domain, courtesy of the U.S Geological Survey (<https://earthexplorer.usgs.gov>)

```{r 03-trypillia, fig.cap="(ref:03-trypillia)", fig.scap="Map with location of Trypillia sites"}
include_graphics("Results/fig03_trypillia.pdf")
```

### Trypillia house construction, architecture and social organisation

As with Linear Pottery houses, Trypillia domestic architecture has been amply documented through more than a century of numerous excavations. However, there are also three additional factors that have contributed to an even more detailed understanding of these latter houses:

-   The very common practice in Trypillia settlements of burning houses entirely at the end of their use life

-   Better post-depositional preservation than most comparable West-European soils, due to less intensive use of modern agro-industrial machinery, leaving architectural remains including living floors most of the time *in situ*, and

-   Numerous finds of detailed miniature clay models of Trypillia houses

Combined, these favourable conditions allow for a relatively precise archaeological reconstruction of how people built their houses and lived in them within the Trypillia culture [see @chernovol2012 for a detailed overview in English]. The main feature of house remains -- which is usually very visible on geomagnetic images -- is a compact clay floor of about 5 to 12 cm thickness interpreted as the main living space of the inhabitants. Imprints of wooden beams are regularly found on the downward side of floor fragments, and it is widely accepted that these floors were constructed on a raised platform of some height, leaving an open space below possibly for storage or even housing for livestock. This configuration is also regularly seen on clay house models, which also hint at the use of rounded ceilings and gables. Walls were made mainly of wattle-and-daub, also seen in wattle imprints on fired daub fragments, but there is also occasional evidence of walls made from horizontal wooden beams -- possibly this technique was applied for wall foundations [@chernovol2012, p. 183]. The main living floor on the raised platform was presumably reached by a ladder from the outside, since openings are not found on the platform floors.

The interior organisation of rooms and installations was very homogeneous across the Trypillia complex, to the degree that Chernovol speculates that there were taboos associated with deviating from the strict norm [-@chernovol2012, p. 200]. For settlements in the Tomashovka local group (see above), the vast majority of houses were rectangular, about 3 to 4 times longer than wide, with the upper level composed of an entrance porch on one end, leading through a doorway to the main room. There, a clay oven was located immediately to the right of the entrance, while a trough and long podium followed the left side wall. By the short end opposite the entrance, there was a special installation usually interpreted as a domestic altar -- a clay disc about 1.6--2m in diameter and 10--15cm thick, often decorated with red and yellow paint and linear incised patterns. Tableware is regularly found at different parts of the house during excavation, notably near the altar and entrance, and the question of whether items were placed there in a meaningful way prior to house burning has been a matter of much discussion [@chernovol2012, p. 198]. Some more rare variations on this rather strict norm of house organisation include secondary altars either in the main room or the porch, as well as an additional annex -- possibly for craft activities -- appended to the front porch.

-   Artisan specialisation in fine ceramics? Metal finds -- gold and copper. @korvin-piotrovskiy2016, add pages.

-   Loom weights are generally found within regular houses, most often on the raised platform near the altar, suggesting that weaving was a domestic activity. This is not the case for remains associated with pottery making and flint knapping, which must have taken place at least outside regular domestic areas, and possibly within specialised workshops, some cases of which have been suggested within CTCC settlements [@chernovol2012, pp. 198-200].

    -   Burning, experimental [@korvin-piotrovskiy2012; @johnston2020]. Collapse and deposit formation.

-   @pickartz2019, pp. 1649-50: "unburnt" and "eroded" are misleading labels, since highly burnt and even vitrified material regularly found during excavation of these. Probably reflecting houses that are constructed with less daub, without an oven installation, and/or without raised clay platform, thus causing significantly weaker magnetic signals. These houses may in fact represent entirely different social functions from regular houses, but there is as for now no consensus as to what this may be (e.g. artisanal, storage etc.)

A special category of much larger houses, generally referred to as "mega-structures" and primarily known from the "mega-sites", was first discovered in 2009 through geomagnetic imagery, and has received much attention since [comparative overview in @hofmann2019; see also @nebbia2018; @chapman2016; @rassmann2016; @burdo2016; @muller2016a; @chapman2014; @rassmann2014]. These are recognised by a number of architectural features that clearly distinguish them from the regular domestic houses described above [@hofmann2019, p. 35 ff.]. Firstly, they usually have twice the floor area or more compared to regular houses -- from about 100m^2^ up to 1.200m^2^ for the largest known examples (in Dobrovody and Nebelivka) -- and are spatially distributed in characteristic ways within the settlements. They are not constructed on raised platforms, and in many cases they are only partially if at all roofed (i.e. with a large walled front yard or atrium). Only a few of these buildings have been excavated -- @hofmann2019 enlist four: one each in Maidanetske, Nebelivka and Dobrovody (all representing Trypillia B-II/C-I mega-sites in the Syniukha basin, central Ukraine) as well as one in Baia (Precucuteni, north-east Romania). The excavated mega-structures in Dobrovody and Nebelivka both represented the largest structures within their settlement, and while the former was interpreted as an entirely unroofed enclosure, several key aspects of the internal architecture of the latter -- the existence of an upper floor, supporting vertical posts and galleries, number of use phases, burning events -- have been subject to highly conflicting views [@gaydarska2020b; @chapman2016; @burdo2016; @chapman2014]. The excavated mega-structure at Maidanetske was not the largest at the site, but by its size and positioning clearly distinguished from domestic houses. It was interpreted by the excavators as being sectioned in two almost equally sized parts, one roofed and one walled yard.

All of the three excavated mega-structures in the Syniukha basin yielded rather regular domestic purpose finds, but in much lower quantities than what would be expected from their sizes if they were permanently inhabited. This, along with the absence of installed ovens, has led the involved research teams to interpret them as being non-domestic in function, i.e. that they did not serve as the homes of ruling-class elites. However, what exactly *did* take place there remains a more open question. The Nebelivka mega-structure housed as much as seven cruciform clay platforms analogous to the "altars" of domestic houses, but some of which were much larger. The Ukrainian part of the excavation team at Nebelivka (see below) have advocated an interpretation of a two-storey building, with the inner room of the upper floor resembling a dramatically up-scaled version of a domestic house, suggesting that this part of the structure could be well suited for permanent occupation [@burdo2016, p. 111], while the mentioned altar platforms -- showing clear traces of fire-related use -- would be located on the lower floor. The authors do not hesitate to characterise the structure as a whole as a communal temple. The British part of the Nebelivka excavation team has favoured an interpretation where only the middle part of the structure would be roofed, and where no rooms would be suited for year-round occupation. While they are more prudent regarding the interpretation of the raised clay platforms as altars (they prefer the more neutral term "Raised Area"), they in turn do not hesitate to designate the complex as an "Assembly House", the term being borrowed by #Whom? US somewhere?

Special finds in Nebelivka mega-structure: @gaydarska2019, pp. 107-9 (graphite, import from Gumeln..)

Among these, a certain size hierarchy is observed in the largest settlements, where size categories are seen to correlate with the placement of the building within the settlement.

-   Social organisation

    -   Lacking burials (but further south?)

    -   Mega-sites as exception,

        -   Urban or not? Density and diversification. Comparative. @diachenko2017; @kruts2012; @gaydarska2019; @muller2016a . Mega-sites as *anomalocivitates*: @ohlrau2022 and @sindbæk2022. Comparison with Uruk: @muller2016b (add pages).

        -   The seasonality hypothesis -- relations to the hinterland: @gaydarska2019, p. 110

        -   Government system -- assemblies or chiefs/kings? @hofmann2019; @chapman2016; @nebbia2018

"*The lack of monumental architecture and the paucity of prestige goods, together with an undeveloped mortuary domain in which to display differences in social status, constitute the main arguments against a permanent, ranked socio-political organization for the Trypillia population. More likely, an inter-regional decision-making political body developed during collective gathering events for the seasonal megasite co-ordination of a generally egalitarian society*" [@nebbia2018, p. 51]. My counter-arguments. But don't insist.

### Sampled Trypillia settlements: The Syniukha basin

-   The German-Ukrainian research team [@hofmann2018]

    -   Maidanetske

    -   Moshuriv

    -   Talne 3

-   The British-Ukrainian team

    -   Nebelivka. The mega-structure as a "temple" with "seven altars" [@burdo2016].

## Reading site plans from geomagnetic imagery

Caveats: fill in here.

@rassmann2016; @rassmann2014 (see my article for precise refs.)

@pickartz2021 fds, @pickartz2019

Say where the shapefiles come from here. @muller-scheessel2020; @hale2020; @ohlrau2020 (double-check in my article)

## Synthetic data

And why I'm not (this time) relying on ethnographic data.

Don't go into technicalities here, just the reasoning.
