# Results: Distribution fitting {#results-distfit}

## Synthetic distributions {#distfit-synth}

### Multiplicative process

A multiplicative process is one in which a term is multiplied with itself a number of times, and can be mathematically described through an exponential function. This is commonly denoted $$f(x) = ab^x$$ read as "the function of *x* equals *a* times *b* to the power of *x*", where *b* is the main term which is multiplied with itself *x* number of times, and *a* being a constant of proportionality. In such functions, *b* determines the *shape* of the function -- if between 0 and 1, the function models exponential decrease or decay as *x* grows larger, and if $b > 1$ the function models exponential growth. This term corresponds to the *rate* parameter for exponential distributions detailed in Section \@ref(distributions). The *a* term determines the *scale* of the function, similarly to the y-intercept of a linear function of the type $f(x) = ax+b$ (in this case *b*), and though in an exponential function this parameter is multiplied rather than added to the variable term (*b^x^*), its value gives the y-intercept since $ab^x = a$ when $x = 0$. Furthermore, the link between exponential and linear functions is illustrated more clearly when we log-transform the former, that is: $f(x) = log(b)x+a$, which is essentially a linear function where *a* is the y-intercept and *b* is the slope (\# Show with fig! and go on).

#Insert figure, ill. exponential functions.

### Additive process

fdsa

Lumping of samples may be done for two main reasons (in archaeology): 1 - the sample size is too small, and lumping together several contemporary samples can raise the sample size to an acceptable level (spatial lumping), or 2 - it is impossible to temporally disentangle elements within a settlement, i.e. reduce temporal resolution to coeval elements, in which case one has no option but temporal lumping (accepting a low temporal resolution). The problem can be further broken down to two case types: a) all lumped samples are really drawn from the same distribution, and b) they are not similarly distributed (**illustrate with table**). For spatial lumping this degree of similarity can be assessed, but not necessarily for temporal lumping. But even when it can be assessed, the lumping needs to be justified in social terms. As an example, several spatial samples (e.g. villages in a region) can have house-size distributions with no significant differences between them (ANOVA), but at the same time be functionally entirely independent, in which case it is more logical to augment the sample size of a single one through simulation and evaluate plausibility through bootstrapping, rather than by lumping of all samples, **rewrite this more clearly**. On the other hand, house sizes can be significantly different between quarters or suburbs within a city or metropolitan area, or even between cities in a region or country, but if they all function together in a coherent system, they may reflect a spatial segregation between different strata in the society, and it makes much sense to lump and analyse them together.

When it comes to temporal lumping, given that a settlement does not undergo substantial cultural changes during its timespan, a workaround to evaluate whether the house-size distribution evolves significantly over time may be to target a number of size categories for ^14^C dating, and to check that they all stretch over the entire range of the settlement's duration, and if yes, accept to analyse the whole distribution as one. **recheck this after I have my results, maybe move it all to summary**. 洧랝

## Settlements

Running the distribution-fitting algorithm presented in Chapter \@ref(methods-distfit) on the house-size distributions of the 13 settlements in the current sample (settlements with 10 or less houses were excluded), resulted in five settlements being interpreted as having power-law tails. These were the Trypillia sites at Maidanetske, Nebelivka and Moshuriv, as well as the Linear Pottery sites at Vr치ble and Horn칳 (Figure \@ref(fig:06-settle-box) and Table \@ref(tab:06-settle-tab)). The remaining eight settlements were interpreted as having exponential tails. It is important to note however, that this simply means that the exponential model was the best out of the tested models, on the tail length that gave the best possible power-law fit. In other words, for these eight settlements, the power-law model is effectively excluded, since other models better explain whatever tail could be interpreted as a power law. Analysing the entire distributions the same way without setting a lower threshold value ($x_{min}$) resulted in all settlements but one being interpreted as having log-normally distributed house sizes. Only 칔쬬ny (Linear Pottery) was interpreted as normally distributed (Table \@ref(tab:06-settle-tab)). This result can also be suspected from the logarithmic box-plot, where most distributions are near symmetric (Figure \@ref(fig:06-settle-box)).

(ref:06-settle-box) House sizes of the 13 analysed settlements, arranged according to median house size. Red dots represent houses with size $\geq x_{min}$ within distribution tails interpreted as power laws. X axis is logarithmic

```{r 06-settle-box, fig.cap="(ref:06-settle-box)"}
load("Results/fig06_settle_box.RData") 
fig06_settle_box
```

Tails.

As shown in Figure \@ref(fig:06-settle-tails)a, with the exception of Horn칳 (Linear Pottery), the power-law tails are much longer than those of the other settlements (b). Even though they do not span two orders of magnitude in size -- as predicted earlier due to material and physical limitations houses as well as the symbolic nature of house size as expression of hierarchy (Section \@ref(power-law)) -- they do so on the cCDF which is based on house rank. Two orders of magnitude in rank represents a decreasing proportion of data points as total number increases, meaning that such power-law tails are "easier" to achieve for larger settlements, which is already an indication that this behaviour is inherently connected to settlement size. Table \@ref(tab:06-settle-tab) shows that the power-law settlements are indeed the largest ones in the sample, again with the exception of Horn칳. While these results are not at all surprising when it comes to the two Trypillia mega-sites -- indeed the hierarchical scaling of buildings there is easily recognised without quantitative analysis -- there are two results that may come as more surprising.

Firstly there is a clearly recognised power law in the house-size distribution of Vr치ble, indicating the probable presence of a hierarchical scaling relationship in its buildings. This is not obvious simply from looking at the site plan, and relates furthermore to the question of social organisation and hierarchy which is much more open and unsettled in Linear Pottery than in Trypillia research. The power-law tail of Horn칳 should be treated with more caution, as it seems more likely to result erroneously from the very small sample size. Secondly -- though the total sample of settlements is admittedly small -- the single most relevant predictor for the presence of scaling in house sizes seems not to be cultural belonging but rather settlement size, since the largest Linear Pottery settlement shows scaling while the smallest Trypillia settlement (Talne 3) does not. This is an indication that social organisation should not be considered as uniform within even well defined archaeological cultures and regions, despite common practice in archaeology. Here, in the case of near contemporary settlements with shared material culture located only some kilometres apart there are two examples of settlements with deviating size that also deviate in how their houses scale internally, possibly indicating very different structures of social organisation between them and the other settlements in their respective regions (hierarchy seen in Vr치ble and not the rest of the 콯itava valley, and equality seen in Talne 3 and not at other sites in the Bug-Dnieper interfluve). It should be worthwhile to test whether this relationship -- between scaling in house sizes and settlement size -- cross-cut archaeological cultures on larger samples.

(ref:06-settle-tails) Analysed house-size distributions for whole settlements. a) Distributions with identified power-law tails. For clarity, only the tails (house sizes $\geq x_{min}$) are coloured and connecting lines are added within each settlement. The grey frame represents the extent of panel b. b) House-size distributions without power-law tails. Two settlements are atypical with their absence of large houses. Scales are logarithmic

```{r 06-settle-tails, fig.cap="(ref:06-settle-tails)"}
load("Results/fig06_settle_tails.RData")
fig06_settle_tails
```

Figure \@ref(fig:06-settle-tails)b shows that the house-size distributions of both Talne 3 (Trypillia) and 칔쬬ny (Linear Pottery) are clearly different than those of the other settlements without power-law tails. At the same time, only 칔쬬ny was interpreted as normally distributed -- i.e. not skewed but symmetric, with only random differences between houses -- when analysed formally. Talne 3 houses were interpreted as a log-normal distribution, but with a very low standard deviation. A Shapiro-Wilk normality test of the house-sizes of Talne 3 gives a test statistic $W = 0.958$ and $p = 0.541$, which is far from enough to exclude the null-hypothesis of a normal distribution. In other words, for the practical purpose here of interpreting probable underlying mechanisms of house-size differences, there is no significant difference between houses at Talne 3 and the distribution can be considered normal.

(ref:06-settle-tab) Main results of distribution fitting analysis on settlements, ordered according to retained tail model (Tail) and parameter values (T_Par1), from steepest to shallowest (see Figure \@ref(fig:06-settle-tails)). Model and Gini are evaluated on the entire distribution for each settlement. Par1 and Par2 represent $\mu$ (mean) and $\sigma$ (standard deviation) for normal and log-normal distributions, and T_Par1 represents $\lambda$ (rate) for exponential and $\alpha$ (scaling exponent) for power-law distributions. $x_{min}$ is the house size from which the tail parameters are estimated. N is the number of houses per settlement, N_tail is the number of houses in the distribution tail, and Tail_Pr is the proportion of N being part of the tail, or N_tail/N

```{r 06-settle-tab}
load("Results/tab06_settle.RData")
kable(tab06_settle, booktabs = TRUE, caption = "(ref:06-settle-tab)") %>% 
#      linesep = c(rep("", 7), "\\addlinespace")) %>%
  kable_styling(latex_options = c("striped", "scale down")) %>% 
  landscape()
```

-   Location on the settlement plans pl-distributed houses (at least Vr치ble and Nebelivka, make in QGIS)

-   Make the point that none of this matters if the power laws are really just artefacts of temporal palimpsest and go on.

\FloatBarrier

## Quarters/neighbourhoods

House-size distributions of separate quarters (Nebelivka) and neighbourhoods (Vr치ble). Can I include separate Nebelivka neighbourhoods as well? (prob. too small, but check)

(ref:06-quart-box) fdsasdf

```{r 06-quart-box, fig.cap="(ref:06-quart-box)"}
load("Results/fig06_quart_box.RData")
fig06_quart_box
```

fsdasdfasfds

(ref:06-quart-tails) fdsasdf

```{r 06-quart-tails, fig.cap="(ref:06-quart-tails)"}
load("Results/fig06_quart_tails.RData")
fig06_quart_tails
```

fdsasdfashgfdghd

(ref:06-quart-tab) fdsasdf

```{r 06-quart-tab}
load("Results/tab06_quart.RData")
kable(tab06_quart, booktabs = TRUE, caption = "(ref:06-quart-tab)") %>% 
  kable_styling(latex_options = c("striped", "scale down")) %>% 
  landscape()
```

fdsasdfasdf

-   Vr치ble, Pareto + site map

-   Nebelivka, Pareto + site map

## Temporal samples (Vr치ble)

If time, I can add more 콯itava sites here, but not necessary. I can also analyse temporal samples of Vr치ble neighbourhoods separately.

(ref:06-time-box) fdsasfjskksl

```{r 06-time-box, fig.cap="(ref:06-time-box)"}
load("Results/fig06_time_box.RData")
fig06_time_box
```

fdsdkklkj칮kj칮lkj

(ref:06-time-tails) fdskjlkj칮lkjuuhkjh

```{r 06-time-tails, fig.cap="(ref:06-time-tails)"}
load("Results/fig06_time_tails.RData")
fig06_time_tails
```

fdsasdfaskjkl

(ref:06-time-tab) Needs to be created!

```{r 06-time-tab}

```

-   Vr치ble, Pareto + site map (exerpts)

-   One neighbourhood, same.

## Summary of findings

END Chapter
