# Results: Distribution fitting {#results-distfit}

## Synthetic distributions {#distfit-synth}

### Multiplicative process

A multiplicative process is one in which a term is multiplied with itself a number of times, and can be mathematically described through an exponential function. This is commonly denoted $$f(x) = ab^x$$ read as "the function of *x* equals *a* times *b* to the power of *x*", where *b* is the main term which is multiplied with itself *x* number of times, and *a* being a constant of proportionality. In such functions, *b* determines the *shape* of the function -- if between 0 and 1, the function models exponential decrease or decay as *x* grows larger, and if $b > 1$ the function models exponential growth. This term corresponds to the *rate* parameter for exponential distributions detailed in Section \@ref(distributions). The *a* term determines the *scale* of the function, similarly to the y-intercept of a linear function of the type $f(x) = ax+b$ (in this case *b*), and though in an exponential function this parameter is multiplied rather than added to the variable term (*b^x^*), its value gives the y-intercept since $ab^x = a$ when $x = 0$. Furthermore, the link between exponential and linear functions is illustrated more clearly when we log-transform the former, that is: $f(x) = log(b)x+a$, which is essentially a linear function where *a* is the y-intercept and *b* is the slope (\# Show with fig! and go on).

#Insert figure, ill. exponential functions.

### Additive process

fdsa

Lumping of samples may be done for two main reasons (in archaeology): 1 - the sample size is too small, and lumping together several contemporary samples can raise the sample size to an acceptable level (spatial lumping), or 2 - it is impossible to temporally disentangle elements within a settlement, i.e. reduce temporal resolution to coeval elements, in which case one has no option but temporal lumping (accepting a low temporal resolution). The problem can be further broken down to two case types: a) all lumped samples are really drawn from the same distribution, and b) they are not similarly distributed (**illustrate with table**). For spatial lumping this degree of similarity can be assessed, but not necessarily for temporal lumping. But even when it can be assessed, the lumping needs to be justified in social terms. As an example, several spatial samples (e.g. villages in a region) can have house-size distributions with no significant differences between them (ANOVA), but at the same time be functionally entirely independent, in which case it is more logical to augment the sample size of a single one through simulation and evaluate plausibility through bootstrapping, rather than by lumping of all samples, **rewrite this more clearly**. On the other hand, house sizes can be significantly different between quarters or suburbs within a city or metropolitan area, or even between cities in a region or country, but if they all function together in a coherent system, they may reflect a spatial segregation between different strata in the society, and it makes much sense to lump and analyse them together.

When it comes to temporal lumping, given that a settlement does not undergo substantial cultural changes during its timespan, a workaround to evaluate whether the house-size distribution evolves significantly over time may be to target a number of size categories for ^14^C dating, and to check that they all stretch over the entire range of the settlement's duration, and if yes, accept to analyse the whole distribution as one. **recheck this after I have my results, maybe move it all to summary**. 洧랝

## Settlements

Running the distribution-fitting algorithm presented in Chapter \@ref(methods-distfit) on the house-size distributions of the 13 settlements in the current sample (settlements with 10 or less houses were excluded), resulted in five settlements being interpreted as having power-law tails. These were the Trypillia sites at Maidanetske, Nebelivka and Moshuriv, as well as the Linear Pottery sites at Vr치ble and Horn칳 (Figure \@ref(fig:06-settle-box) and Table \@ref(tab:06-settle-tab)). The remaining eight settlements were interpreted as having exponential tails. It is important to note however, that this simply means that the exponential model was the best out of the tested models, on the tail length that gave the best possible power-law fit. In other words, for these eight settlements, the power-law model is effectively excluded, since other models better explain whatever tail could be interpreted as a power law. Analysing the entire distributions the same way without setting a lower threshold value ($x_{min}$) resulted in all settlements but one being interpreted as having log-normally distributed house sizes. Only 칔쬬ny (Linear Pottery) was interpreted as normally distributed (Table \@ref(tab:06-settle-tab)). This result can also be suspected from the logarithmic box-plot, where most distributions are near symmetric (Figure \@ref(fig:06-settle-box)).

(ref:06-settle-box) House sizes of the 13 analysed settlements, arranged according to median house size. Red dots represent houses with size $\geq x_{min}$ within distribution tails interpreted as power laws. X axis is logarithmic

```{r 06-settle-box, fig.cap="(ref:06-settle-box)"}
load("Results/fig06_settle_box.RData") 
fig06_settle_box
```

As shown in Figure \@ref(fig:06-settle-tails)a, with the exception of Horn칳 (Linear Pottery), the power-law tails are much longer than those of the other settlements (b). Even though they do not span two orders of magnitude in size -- as predicted earlier due to material and physical limitations of houses as well as the symbolic nature of house size as expression of hierarchy (Section \@ref(power-law)) -- they do so on the cCDF which is based on house rank. Two orders of magnitude in rank represents a decreasing proportion of data points as total number increases, meaning that such power-law tails are "easier" to achieve for larger settlements, which is already an indication that this behaviour is inherently connected to settlement size. Table \@ref(tab:06-settle-tab) shows that the power-law settlements are indeed the largest ones in the sample, again with the exception of Horn칳. While these results are not at all surprising when it comes to the two Trypillia mega-sites -- indeed the hierarchical scaling of buildings there is already recognised without quantitative analysis -- there are two results that may come as more surprising.

Firstly there is a clearly recognised power law in the house-size distribution of Vr치ble, indicating the probable presence of a hierarchical scaling relationship in its buildings. This is not obvious simply from looking at the site plan, and relates furthermore to the question of social organisation and hierarchy which is much more open and unsettled in Linear Pottery than in Trypillia research. The power-law tail of Horn칳 should be treated with more caution, as it seems more likely to result erroneously from the very small sample size. Secondly -- though the total sample of settlements is admittedly small -- the single most relevant predictor for the presence of scaling in house sizes seems not to be cultural attribution but rather settlement size, since the largest Linear Pottery settlement shows scaling while the smallest Trypillia settlement (Talne 3) does not. This is an indication that social organisation should not be considered as uniform within even well defined archaeological cultures and regions, despite common practice in archaeology. Here, in the case of near contemporary settlements with shared material culture located only some kilometres apart, there are two examples of settlements with deviating size that also deviate in how their houses scale internally, possibly indicating very different structures of social organisation between them and the other settlements in their respective regions (hierarchy seen in Vr치ble and not the rest of the 콯itava valley, and equality seen in Talne 3 and not at other sites in the Bug-Dnieper interfluve). It should be worthwhile to test whether this relationship -- between scaling in house sizes and settlement size -- also cross-cut archaeological cultures on larger samples.

(ref:06-settle-tails) Analysed house-size distributions for whole settlements. a) Distributions with identified power-law tails. For clarity, only the tails (house sizes $\geq x_{min}$) are coloured and connecting lines are added within each settlement. The grey frame represents the extent of panel b. b) House-size distributions without power-law tails. Two settlements are atypical with their absence of large houses. Scales are logarithmic

```{r 06-settle-tails, fig.cap="(ref:06-settle-tails)"}
load("Results/fig06_settle_tails.RData")
fig06_settle_tails
```

Figure \@ref(fig:06-settle-tails)b shows that the house-size distributions of both Talne 3 (Trypillia) and 칔쬬ny (Linear Pottery) are clearly different from those of the other settlements without power-law tails. At the same time, only 칔쬬ny was interpreted as normally distributed -- i.e. not skewed but symmetric, with only random differences between houses -- when analysed formally. Talne 3 houses were interpreted as a log-normal distribution, but with a very low standard deviation. A Shapiro-Wilk normality test of the house sizes of Talne 3 gives a test statistic $W = 0.958$ and $p = 0.541$, which is far from enough to exclude the null-hypothesis of a normal distribution. In other words, for the practical purpose here of interpreting probable underlying mechanisms of house-size differences, there is no significant difference between houses at Talne 3 and the distribution can be considered normal.

(ref:06-settle-tab) Main results of distribution fitting analysis on settlements, ordered according to retained tail model (Tail) and parameter values (T_Par1), from steepest to shallowest (see Figure \@ref(fig:06-settle-tails)). Model and Gini are evaluated on the entire distribution for each settlement. Par1 and Par2 represent $\mu$ (mean) and $\sigma$ (standard deviation) for normal and log-normal distributions, and T_Par1 represents $\lambda$ (rate) for exponential and $\alpha$ (scaling exponent) for power-law distributions. $x_{min}$ is the house size from which the tail parameters are estimated. N is the number of houses per settlement, N_tail is the number of houses in the distribution tail, and Tail_Pr is the proportion of N being part of the tail, or N_tail/N

```{r 06-settle-tab}
load("Results/tab06_settle.RData")
kable(tab06_settle, booktabs = TRUE, caption = "(ref:06-settle-tab)") %>% 
#      linesep = c(rep("", 7), "\\addlinespace")) %>%
  kable_styling(latex_options = c("striped", "scale down")) %>% 
  landscape()
```

From the table it is clear that the Gini index follows the model type and parameter values closely but not exactly. The advantage of using the Gini index for the purpose of quantifying inequality is that it is more straight-forward to calculate, and it allows for a unified single measure facilitating comparison between different distribution types [e.g. @kohler2018; @kohler2017]. The disadvantage is that there is no clear and reliable way of identifying underlying mechanisms from this index alone -- in the present case it would be impossible to single out which settlements show signs of hierarchical scaling in their houses, let alone determining how many of the houses this would concern. My conclusion from this is that the Gini index is useful when the question to be answered relates specifically to inequality within populations, and when cross-cultural comparisons or long temporal trends are more emphasised than the social organisation of specific cultural contexts.

Looking at how the power-law distributed houses are spread spatially within the various settlements allows for further interesting patterns to emerge. At Vr치ble, the whole range of hierarchically scaling houses is present at each of the neighbourhoods, though apparently somewhat less in the south-west (Figure \@ref(fig:06-vrable-map)). There seems to be a weak tendency for larger houses to cluster throughout, i.e. that large houses tend to be close to other large houses, and with interstices between clusters being filled by regular small houses of size below $x_{min}$. However, observing the entire settlement plan in a single block does not allow for further evaluation of whether these clusters actually represent coeval groups of larger houses or if they rather form temporal sequences, following the much discussed "yard model" of Linear Pottery settlement development [@zimmermann2012; \@ref(lbk)]. Either option speaks against an interpretation of house-size differences being only random.

![Power-law distributed houses at Vr치ble (Linear Pottery), arbitrarily grouped to three levels using the Jenks optimisation method integrated in QGIS for readability. Figure made by author with data from @muller-scheessel2020](Results/fig06_vrable_pl.pdf){#fig:06-vrable-map}

In the much larger Trypillia settlements of Maidanetske (Figure \@ref(fig:06-maidanetske-map)) and Nebelivka (Figure \@ref(fig:06-nebelivka-map)) the hierarchical scaling of houses is seen throughout the settlements, indicating that there is no general sector that has obviously more of smaller or larger houses than others. As is widely recognised by the researchers who have studied these settlements in detail [e.g. @hale2020, pp. 127-37; @ohlrau2020, pp. 61-4], the single and by far largest building is located at around 3 o'clock on the inner house ring (at least for Nebelivka, but most probably also Maidanetske), while other smaller mega-structures or "Assembly Houses" are regularly dispersed throughout the main street and, to a lesser extent and mostly for Nebelivka, outside the outer house ring. The most interesting result from the distribution fitting analysis provided here, is that the scaling behaviour also goes far beyond the typological distinction between mega-structures and domestic houses. For both settlements, extra large domestic houses are more concentrated around the main street than in the inner streets and plazas, and while these are quite regularly spaced in Maidanetske, in Nebelivka they seem to cluster near the Assembly Houses, further underlining the size hierarchy. But again it is difficult to know at the present stage to which extent these clusters represent coeval groups of larger houses or temporal sequences, given the duration of the settlement. It does however suggest in any case, that large domestic houses had privileged access to the Assembly Houses by close proximity, while the smaller houses were mostly confined to the inner streets and the borders between Quarters, i.e. the furthest away from the Assembly Houses.

![Power-law distributed houses at Maidanetske (Trypillia), grouped to three levels with Jenks optimisation. The levels are arbitrary but overlap well with the typological distinction between mega-structures (dark blue) and other houses. Hierarchical scaling includes far more houses than the mega-structures, and is distributed across the settlement. Figure made by author with data from @ohlrau2020](Results/fig06_maidanetske_pl.pdf){#fig:06-maidanetske-map}

![Power-law distributed houses at Nebelivka (Trypillia), grouped into four arbitrary levels with Jenks optimisation, the two largest of which are overlapping with the typological levels of the "Mega-structure" and the "Assembly Houses". Figure made by author with data from @hale2020](Results/fig06_nebelivka_pl.pdf){#fig:06-nebelivka-map}

In the small Trypillia settlement of Moshuriv, power-law distributed houses are also spread across the settlement plan, with the largest house situated in the east end of the inner house ring, following the same scheme as the mega-sites, as noted by Ohlrau [-@ohlrau2020, pp. 241-3]. Here, much like at Maidanetske in miniature, the hierarchically scaling houses are evenly spread without obvious clustering, the only pattern being that the three houses just below the largest one in size also are located on the inner house ring (Figure \@ref(fig:06-moshuriv-map)).

![Power-law distributed houses at Moshuriv (Trypillia), grouped arbitrarily into three size categories using Jenks optimisation. Made by author with data from @ohlrau2020](Results/fig06_moshuriv_pl.pdf){#fig:06-moshuriv-map}

Since the main goal in this thesis is not only to measure and interpret archaeological data, but also to assess the reliability of the methods and the robustness of the results, in the following sections I propose further iterations of this analysis on gradually more filtered sub-sets of the data, to see whether the power-law interpretations continue to hold. Various factors specifically related to archaeological data could be thought to influence size distributions in ways that would preclude any meaningful interpretations in terms of social organisation. More specifically, if power-law distributed house sizes are to be considered as statistical signatures of social hierarchy, the possibility of them being no more than artefacts resulting from methodological choices like spatial or temporal lumping needs to be excluded.

\FloatBarrier

## Quarters/neighbourhoods

Two of the large settlements with power-law house-size distribution tails -- Nebelivka and Vr치ble -- were selected for further distribution fitting analysis within separate quarters or neighbourhoods. The Linear Pottery site of Vr치ble is clearly organised into three distinct neighbourhoods, termed North (N), South-West (SW) and South-East (SE) by the research team in the dedicated publications [@furholt2020a; @winkelmann2020]. The Trypillia mega-site of Nebelivka has also been subdivided into quarters by its research team (following their terminology here), but the limits between them were set somewhat more arbitrarily based on a series of criteria, including natural topography, placement and orientation of streets and entrance ways, as well as the locations of the large so-called Assembly Houses [@hale2020, p. 123]. This resulted in a series of 14 defined quarters arbitrarily labelled A--N, which is reused as is here. It should be noted that the single largest building in Nebelivka -- the "Mega-structure" of approximately 1.200 m^2^ -- is not included in any of these quarters, but lies in the interstice between quarters A and B along with a few other houses. Though this delimitation of quarters at Nebelivka could have been done differently, which is clearly recognised by the researchers themselves, I do not pretend that I could make any better judgement. The number of houses per quarter is included in Table \@ref(tab:06-quart-tab) below.

(ref:06-quart-box) House-size distributions of individual quarters for Nebelivka and neighbourhoods for Vr치ble, arranged according to median house size. Red dots indicate houses of size $\geq x_{min}$ in cases where the distribution tail was interpreted as a power law. X axis is logarithmic

```{r 06-quart-box, fig.cap="(ref:06-quart-box)"}
load("Results/fig06_quart_box.RData")
fig06_quart_box
```

When analysed separately following the same distribution-fitting algorithm as with the entire settlements, 14 of the 17 data series were recognised as having power-law tails (Figures \@ref(fig:06-quart-box) and \@ref(fig:06-quart-tails)). The only deviations were Nebelivka quarters E and N, and the SE neighbourhood of Vr치ble. This clearly shows that the power laws observed at the total settlement level do not merely result from analytically stacking together separate more lightly skewed distributions like log-normals, but rather from genuinely different scaling patterns in house sizes for these settlements. The $\alpha$ estimates, between 3.6 and 6.8 (with one outlier for Neb. A), were largely similar to the ones observed for the entire settlements (5.2 for Vr치ble and 4.8 for Nebelivka; Tables \@ref(tab:06-quart-tab) and \@ref(tab:06-settle-tab)). Each of the quarters and neighbourhoods were furthermore identified as log-normal when analysed over their entire size range, with $\mu$ and $\sigma$ estimates very similar to those observed for the total settlements.

(ref:06-quart-tails) Survival function of the same house-size distributions of quarters/neighbourhoods. Coloured dots represent houses within power-law tails. Three series were not interpreted as power laws. Scales are logarithmic

```{r 06-quart-tails, fig.cap="(ref:06-quart-tails)"}
load("Results/fig06_quart_tails.RData")
fig06_quart_tails
```

(ref:06-quart-tab) Results of distribution fitting on separate quarters at Nebelivka and neighbourhoods at Vr치ble, arranged by tail model and parameter. Par1 and Par2 indicate $\mu$ and $\sigma$ for log-normal distributions, and T_Par1 is $\alpha$ for power-law and $\lambda$ for exponential tail distributions. Gini index is calculated on the entire sample

```{r 06-quart-tab}
load("Results/tab06_quart.RData")
kable(tab06_quart, booktabs = TRUE, caption = "(ref:06-quart-tab)") %>% 
  kable_styling(latex_options = c("striped", "scale down")) %>% 
  landscape()
```

In addition to testing whether the power-law interpretations hold when the analysis is done at the quarter rather than settlement level, performing distribution fitting on separate quarters allows for observing intra-site differences between quarters. The immediate interpretation of the two non-power law quarters at Nebelivka would be that their houses are more similar in size and that this could indicate a flatter hierarchical structure, or an absence of hierarchy altogether. The association between settlement size and power-law distributions noted above could also be valid here. With only 50 houses, quarter E is the smallest of the Nebelivka quarters (Table \@ref(tab:06-quart-tab), Figure \@ref(fig:06-quart-map)). With a smaller population the inhabitants could possibly do well without a long-range hierarchy beyond the (comparatively small) assembly house located within its section of the central street. However, this interpretation does not fit for quarter N which is in the mid-range of house count with 111 houses. Here the explanation could be taphonomical, as the quarter is lacking an Assembly House which could be located in an unpreserved part of the site [@hale2020, p. 123]. Interestingly, quarter A also has a small house count (51) *and* is lacking an assembly house, but its house-size distribution is all the same interpreted as a power law, albeit with atypical parameter values -- $x_{min}$ is the highest among the Nebelivka quarters, and $\alpha$ is by far the highest -- indicating that these factors may influence model selection without determining it entirely. For quarter E in particular, another possibility which is linked at least to house count, is that its quarter borders are drawn incorrectly so that the quarter should either be larger or included into one of the neighbouring quarters, which would then probably still be power-law distributed. The explanations based on taphonomy and quarter border definition work under the assumption that since most of the quarters have power-law distributed house sizes, all of them should originally have had this, and that the data is somehow distorted. The social explanation of differing degrees of hierarchical organisation between quarters remains a possibility, but at this stage any of these explanations are difficult to exclude. Given that the Nebelivka mega-site is exceptionally well preserved for its size -- the site plan is very nearly complete -- this illustrates the difficulty archaeologists face when trying to interpret only partially preserved settlements. I personally see the Nebelivka plan as extremely regular between quarters, and the mentioned deviations in quarters N and A as resulting from taphonomy (missing data) and in quarter E as incorrectly drawn quarter borders. I also question the border drawn between quarters J and K, though this seemingly has not influenced their internal house-size distributions noticeably.

![Power-law distributed houses in quarters D to G in Nebelivka, fitted by quarter (top) and for the settlement as a whole (bottom). Quarter-wise distribution fitting does not identify hierarchical scaling in quarter E, though many of the houses there are included in the power-law model for the whole settlement. Size categories are arbitrary (three levels with Jenks optimisation) and values differ between quarters. Legend values correspond to quarter F. Figure by author with data from @hale2020](Results/fig06_quart_map.pdf){#fig:06-quart-map}

Regarding Vr치ble SE, in addition to not having a power-law tail, the house-size distribution is also distinguished from the two others with a markedly higher median, i.e. generally larger houses, as well as higher mean and standard deviation (Table \@ref(tab:06-quart-tab)). On the survival function plot on Figure \@ref(fig:06-quart-tails) this translates to a wider and more regular parabolic curve of house sizes, more typically log-normal, as opposed to the more abruptly descending straight lines of the two other neighbourhoods (in red on the plot). For the N and SW neighbourhoods, the combination of a less skewed log-normal for the main body of the distribution (i.e. regular houses) and an actual power law for the largest 41 and 20 houses respectively (1/3 and 1/5 of the houses), could indicate a more marked difference between these, which could be interpreted either as an emerging social elite or the presence of some large houses with special social functions. However, any further interpretation in social terms depends on how many of these houses were actually coeval, i.e. in use at the same time. If most of these large and hierarchically scaling houses were in reality spread out through time, so that only a very few existed at any single one moment, it would not make much sense to speak of a social hierarchy, and the distribution should not be interpreted as a power law. For interpreting social relationships between different households and how they function in daily life, it is essential that the studied households be contemporary -- a requirement that is often hard to meet in archaeology, but an attempt is done in the following for Vr치ble only.

\FloatBarrier

## Temporal samples (Vr치ble)

Separating an archaeological settlement plan into samples with temporally coeval structures only is a demanding exercise, and more so the larger the settlement. The best and most reliable method is to date every single structure in the settlement, but this is usually impossible both in theory, since not all structures yield datable material, and in practice when considering the costs involved. The more realistic approach is to sample a smaller amount of structures throughout the settlement for dating and modelling, and then extrapolate the results to get a more or less rough overview of how much, proportionally, of the site was occupied at the same time. This has been done for the Trypillia mega-sites, indicating that the maxima of coeval habitation would be at approximately 33% of houses at Nebelivka [@millard2020, pp. 253-6; @muller2022] and 52% at Maidanetske [@ohlrau2020, pp. 233-5; @ohlrau2022, pp. 86-8]. Even though some general tendencies of the spatio-temporal development of these settlement plans have been proposed [@shatilo2021, p. 247; @muller2022, pp. 218-9], it is for now impossible to extract precise temporal samples of coeval house-size distributions here. This also goes for the image analysis presented in Chapter \@ref(images-results) [see also @bruvollforthcoming]. The smaller sites of Moshuriv and Talne 3 are only dated through relative chronology of surface finds and thus attributed as a whole to their chronological phase, though their small sizes and ordered layout would suggest that most houses could be coeval [@ohlrau2020, pp. 241-4].

For houses in Linear Pottery settlements, a special dating proxy was recently developed, based on the observation that within this cultural context houses were constructed to be parallel to pre-existing ones, but with a slow and gradual counter-clockwise shift in orientation of approximately 1.3 degrees per decade [@muller-scheessel2020]. This shift was hypothesised to be too slow for any conscious intention among the house builders, but rather a result of so-called pseudoneglect, or the tendency of a slight leftwards bias in the perception of parallel lines, mostly among the right-handed. @muller-scheessel2020 observed this gradual shift in the orientation of 17 houses in Vr치ble for which construction dates had been estimated based on Bayesian modelling of ^14^C dates. Their linear regression model for all these houses only gave a weak correlation however, and they argued for basing the model on the eight houses in the SW neighbourhood only, where the sampling strategy had been the most systematic. This gave a regression model defining house orientation as $0.129x - 651.016$ with correlation coefficient $r = 0.84$, $x$ being the modelled construction year BCE, which when applied inversely (solving for $x$) gives modelled construction year as $\frac{orientation + 651.016}{0.129}$. Use-life of houses in Vr치ble was modelled by @meadows2019 to a median of 27.5 years, a number which according to the same authors has varied greatly throughout the long history of Linear Pottery research. Extrapolating the model for construction year based on house orientation and adding this median duration onto all known houses in Vr치ble gives a settlement occupation span from 5297 to 4975 cal. BCE, which is in good agreement with the currently available and modelled ^14^C dates [though admittedly a few decades earlier, see @meadows2019]. For the distribution fitting analysis in this section I defined a set of 16 sample dates spaced 20 years apart from 5290 to 4990 within this timespan, and assigned houses to each sample accordingly. Since the duration between time samples is a few years shorter than the duration of houses, some houses were assigned to more than one sample. Furthermore, as with the analyses above, samples with 10 or less houses in them were filtered out, mainly since the distribution method then becomes too unreliable (this is already a quite generous allowance, Chapter \@ref(methods-distfit)), but also since it makes little sense to speak of hierarchy in a group of less than 10.

It must be noted that there are several caveats with this procedure of selecting out houses to be considered coeval. Firstly, the dating proxy method proposed by @muller-scheessel2020 is recent and not yet well established. It is based on linear regression of very few data points (8 houses in only one part of the settlement), all of which are not unique certain values but dates that are themselves also modelled. Though the publication does not specify it explicitly, it seems their model was fitted by ordinary least squares (OLS) regression, while systematic measurement uncertainty in the independent variable (the modelled ^14^C dates) should warrant more robust methods like orthogonal regression or probabilistic Monte Carlo methods. My use of a single value for house duration rather than the probability distribution that it really is, is also a simplification that could affect the results of the analysis. In general, longer house durations lead to more of the houses being coeval throughout the timespan of a settlement, while inversely shorter house durations lead to fewer coeval houses. There could also potentially be systematic relationships between house size and duration, e.g. that larger houses were occupied by more temporally stable households for several generations, putting in greater efforts to maintenance. These questions are not further pursued here however, and the results of this analysis must be considered preliminary.

Following this procedure for defining data series with houses in Vr치ble considered to be coeval at given time samples, these series were analysed through the same distribution fitting algorithms as before. Ten of the 16 time samples had more then 10 houses, and these were situated consecutively in time between 5230 and 5050 cal. BCE, the remaining samples being at the start and end of the settlement's duration. Of these 10 samples, seven had tails in their house-size distributions that were interpreted as power laws (Figures \@ref(fig:06-time-box) and \@ref(fig:06-time-tails)).

(ref:06-time-box) House-size distributions at single time samples for the entire settlement at Vr치ble (a) and the South-West neighbourhood only (b). Red dots indicate houses with power-law distributed size. Only distributions including more than 10 houses are represented. In plot b, the single largest house of each sample is also excluded. X axis is logarithmic

```{r 06-time-box, fig.cap="(ref:06-time-box)"}
load("Results/fig06_time_box.RData")
fig06_time_box
```

A last attempt to test if these power laws were simply statistical artefacts of lumped data sets, was done parting from the following hypothesis: At any given time during Vr치ble's duration, the three neighbourhoods were largely independent from each other and there was no hierarchical relationship between them. Within each neighbourhood, household sizes were log-normally distributed, or lightly skewed, as a result of their generalised post-marital residence pattern of patrilocality, and in addition there was always one single building markedly larger than the others, intended for communal ceremonial use. Analysing all houses in Vr치ble together, even by considering only temporally coeval houses, would be enough to generate a false impression of a hierarchy between the largest houses.

In order to investigate if this was the case, only the houses from the south-west neighbourhood of Vr치ble were selected out (being the ones for which the construction date model was the most secure), and the single largest house of each time sample was excluded. It then remained only five time samples with more than 10 houses, namely the ones situated between 5190 and 5110. Out of these, three were still interpreted with power-law distributed house-size distributions (Figures \@ref(fig:06-time-box)b and \@ref(fig:06-time-tails)b). These power laws could thus not be explained as resulting from neither spatial nor temporal lumping, nor from the eventuality of a special function of the single largest house, and it seems therefore reasonable to conclude that they represent actual hierarchical scaling which can be further interpreted in terms of social organisation. It also seems plausible that analysing the three neighbourhoods together may be justified, which allows for identifying this scaling behaviour over a larger range of the settlement's duration. Though conclusions should not be taken too far given the many uncertainties related to the temporal modelling applied here, it is interesting to note that power-law tails are identified throughout the duration of the settlement, and are seemingly not a feature that is specifically related to the later phases only. This seemingly contradicts the current understanding of the temporal trajectory of inequality in Linear Pottery society, which is generally seen as increasing towards the later phases, leading to the rising tensions seen in the widely discussed massacre deposits, including in the very same Vr치ble settlement [@furholt2020b; @muller-scheessel2021; Section \@ref(lbk)]. Not surprisingly, the more conventional inequality measure of the Gini index draws a similar picture of the temporal development of house sizes at Vr치ble (\@ref(tab:06-time-tab)). Throughout the phases with 10 or more coeval houses, the index remains rather stable at values between 0.28 and 0.20, and if anything it shows a more decreasing than increasing trend. In other words, the dynamics of changing house sizes at Vr치ble do not give support to an interpretation of the eventual decline of the settlement as a result of rising inequalities (though this does not exclude the possibility such inequalities being expressed through other material proxies).

(ref:06-time-tails) Survival function of the same house-size distributions of time samples for Vr치ble (a) and Vr치ble SW only excluding the largest house of each sample (b). Coloured dots represent houses within power-law tails. Power-law tails persist despite the gradual breaking down of the data set. Scales are logarithmic

```{r 06-time-tails, fig.cap="(ref:06-time-tails)"}
load("Results/fig06_time_tails.RData")
fig06_time_tails
```

(ref:06-time-tab) Distribution analysis results for Vr치ble, subdivided into time samples with coeval houses, arranged chronologically. The analysis was also done on the South-West neighbourhood separately, where the single largest house for each sample was excluded. In both series, only samples consisting of 10 houses or more were analysed. Par1 and Par2 are $\mu$ and $\sigma$ for log-normal and normal distribution models. T_Par1 is $\alpha$ for power-law and $\lambda$ for exponential tail models. Tail_P is the proportion of data points (N) in the tail model (N_Tail), or N_Tail/N. Gini index is calculated on the entire sample distribution

```{r 06-time-tab}
load("Results/tab06_time.RData")
kable(dplyr::select(tab06_time, 1:11), booktabs = TRUE,
      caption = "(ref:06-time-tab)") %>% 
   kable_styling(latex_options = c("striped", "scale down")) %>% 
   pack_rows(index = c("Vr\u00e1ble" = 10, "Vr\u00e1ble SW*" = 5)) %>% 
   landscape()

```

Three examples of the spatial distribution of coeval power-law houses is seen in Figure \@ref(fig:06-vrable-time-map). Through most of the analysed time samples there is seemingly only one or two very large houses per neighbourhood, and in the first and last of the samples that are large enough for distribution fitting (5230 and 5050 cal. BCE) there is only one for the entire settlement (Figure \@ref(fig:06-time-box)a). This further indicates that it would be in the middle phases of Vr치ble in particular -- roughly between 5150 and 5090 following the model -- that the three neighbourhoods developed into independently functioning settlements, possibly with increased competition between them, as is also suggested by the construction of the enclosure and palisade around the south-west neighbourhood shortly after and by 5070 BCE [@furholt2020a, pp. 493-8]. A single house per phase which is clearly larger than the rest is a regular trend in Linear Pottery settlements, though absolute size categories vary between contexts [@coudart1998, p. 49]. This largest house does not necessarily represent any obvious architectural particularities indicating special function like the Modderman 1a type [-@modderman1970], though they do tend to have double posts (possible elevated granaries) in the frontal section [@coudart1998, pp. 41-2; \@ref(lbk)]. Though this is not yet confirmed through excavation in Vr치ble (#CHECK), the largest houses are generally upscaled versions of smaller common types, in particular the type 1b, or *Gro륵au* with ditch in the posterior section only. And though it is difficult to prove with house sizes alone, it seems fully plausible that such largest houses be inhabited by clan leader households. Once the settlement grew large enough for more than one clan to take on a leading role -- i.e. too large for one clan to control alone -- the settlement would split and crystallise into three more or less independent factions, each with a clan-leading household during the middle phases. Following this model, it would then be the competition between three clan households or Houses that lead to the increased tensions and violence seen in the late phases of the settlement, rather than generally increasing levels of inequality. Given that the above described power-law distributions stretch well beyond these largest houses in the Vr치ble neighbourhoods, it is furthermore tempting to interpret the placement of subsequent houses in the size hierarchy as a function of their kin proximity to the leading household, with the smallest houses (those outside the power law) representing recently established or otherwise poorly integrated households [@hachem2014; \@ref(lbk)].

![Vr치ble at three temporal samples, with house sizes above $x_{min}$ grouped into three arbitrary classes by Jenks optimisation. Parameter values differ slightly between time samples (see table), and the legend categories refer to the 5130 sample. Counter-clockwise shift in house orientation is used as proxy for construction date](Results/fig06_vrable_time.pdf){#fig:06-vrable-time-map}

\FloatBarrier

## Summary of findings

END Chapter
