# Results: Distribution fitting {#results-distfit}

## Synthetic distributions {#distfit-synth}

### Multiplicative process

A multiplicative process is one in which a term is multiplied with itself a number of times, and can be mathematically described through an exponential function. This is commonly denoted $$f(x) = ab^x$$ read as "the function of *x* equals *a* times *b* to the power of *x*", where *b* is the main term which is multiplied with itself *x* number of times, and *a* being a constant of proportionality. In such functions, *b* determines the *shape* of the function -- if between 0 and 1, the function models exponential decrease or decay as *x* grows larger, and if $b > 1$ the function models exponential growth. This term corresponds to the *rate* parameter for exponential distributions detailed in Section \@ref(distributions). The *a* term determines the *scale* of the function, similarly to the y-intercept of a linear function of the type $f(x) = ax+b$ (in this case *b*), and though in an exponential function this parameter is multiplied rather than added to the variable term (*b^x^*), its value gives the y-intercept since $ab^x = a$ when $x = 0$. Furthermore, the link between exponential and linear functions is illustrated more clearly when we log-transform the former, that is: $f(x) = log(b)x+a$, which is essentially a linear function where *a* is the y-intercept and *b* is the slope (\# Show with fig! and go on).

#Insert figure, ill. exponential functions.

### Additive process

fdsa

Lumping of samples may be done for two main reasons (in archaeology): 1 - the sample size is too small, and lumping together several contemporary samples can raise the sample size to an acceptable level (spatial lumping), or 2 - it is impossible to temporally disentangle elements within a settlement, i.e. reduce temporal resolution to coeval elements, in which case one has no option but temporal lumping (accepting a low temporal resolution). The problem can be further broken down to two case types: a) all lumped samples are really drawn from the same distribution, and b) they are not similarly distributed (**illustrate with table**). For spatial lumping this degree of similarity can be assessed, but not necessarily for temporal lumping. But even when it can be assessed, the lumping needs to be justified in social terms. As an example, several spatial samples (e.g. villages in a region) can have house-size distributions with no significant differences between them (ANOVA), but at the same time be functionally entirely independent, in which case it is more logical to augment the sample size of a single one through simulation and evaluate plausibility through bootstrapping, rather than by lumping of all samples, **rewrite this more clearly**. On the other hand, house sizes can be significantly different between quarters or suburbs within a city or metropolitan area, or even between cities in a region or country, but if they all function together in a coherent system, they may reflect a spatial segregation between different strata in the society, and it makes much sense to lump and analyse them together.

When it comes to temporal lumping, given that a settlement does not undergo substantial cultural changes during its timespan, a workaround to evaluate whether the house-size distribution evolves significantly over time may be to target a number of size categories for ^14^C dating, and to check that they all stretch over the entire range of the settlement's duration, and if yes, accept to analyse the whole distribution as one. **recheck this after I have my results, maybe move it all to summary**. 洧랝

## Settlements

Running the distribution-fitting algorithm presented in Chapter \@ref(methods-distfit) on the house-size distributions of the 13 settlements in the current sample (settlements with 10 or less houses were excluded), resulted in five settlements being interpreted as having power-law tails. These were the Trypillia sites at Maidanetske, Nebelivka and Moshuriv, as well as the Linear Pottery sites at Vr치ble and Horn칳 (Figure \@ref(fig:06-settle-box) and Table \@ref(tab:06-settle-tab)). The remaining eight settlements were interpreted as having exponential tails. It is important to note however, that this simply means that the exponential model was the best out of the tested models, on the tail length that gave the best possible power-law fit. In other words, for these settlements, the power-law model is effectively excluded, since other models better explain whatever tail could be interpreted as a power law. The box-plot on Figure \@ref(fig:06-settle-box) indicates that many of the analysed distributions should be well approximated as log-normals, since they show near-symmetric boxes and whiskers on a logarithmic scale. A closer look at the non-power-law distributions reveals that two settlements deviate clearly from the others, namely the Trypillia site at Talne 3 and the Linear Pottery site at 칔쬬ny, both of which are the smallest settlements in their respective culture groups (Figure \@ref(fig:06-settle-tails)).

(ref:06-settle-box) House sizes of the 13 analysed settlements, arranged according to median house size. Red dots represent houses with size $\geq x_{min}$ within distribution tails interpreted as power laws. X axis is logarithmic

```{r 06-settle-box, fig.cap="(ref:06-settle-box)"}
load("Results/fig06_settle_box.RData") 
fig06_settle_box
```

fdsasdfasf

(ref:06-settle-tails) Analysed house-size distributions for whole settlements. a) Distributions with identified power-law tails. For clarity, only the tails (house sizes $\geq x_{min}$) are coloured and connecting lines are added within each settlement. The grey frame represents the extent of panel b. b) House-size distributions without power-law tails. Two settlements are atypical with their absence of large houses. Scales are logarithmic

```{r 06-settle-tails, fig.cap="(ref:06-settle-tails)"}
load("Results/fig06_settle_tails.RData")
fig06_settle_tails
```

fdsasdf

(ref:06-settle-tab) Main results of distribution fitting analysis on settlements, ordered according to retained tail model (Tail) and parameter values (T_Par1), from steepest to shallowest (see Figure \@ref(fig:06-settle-tails)). Model and Gini are evaluated on the entire distribution for each settlement. Par1 and Par2 represent $\mu$ (mean) and $\sigma$ (standard deviation) for normal and log-normal distributions, and T_Par1 represents $\lambda$ (rate) for exponential and $\alpha$ (scaling exponent) for power-law distributions. $x_{min}$ is the house size from which the tail parameters are estimated. N is the number of houses per settlement, N_tail is the number of houses in the distribution tail, and Tail_Pr is the proportion of N being part of the tail, or N_tail/N

```{r 06-settle-tab}
load("Results/tab06_settle.RData")
kable(tab06_settle, booktabs = TRUE, caption = "(ref:06-settle-tab)") %>% 
#      linesep = c(rep("", 7), "\\addlinespace")) %>%
  kable_styling(latex_options = c("striped", "scale down")) %>% 
  landscape()

```

-   Plan of some of the settlements, with coloured pl-distributed houses (at least Vr치ble and Nebelivka, make in QGIS)

## Quarters/neighbourhoods

House-size distributions of separate quarters (Nebelivka) and neighbourhoods (Vr치ble). Can I include separate Nebelivka neighbourhoods as well? (prob. too small, but check)

-   Vr치ble, Pareto + site map

-   Nebelivka, Pareto + site map

## Temporal samples (Vr치ble)

If time, I can add more 콯itava sites here, but not necessary. I can also analyse temporal samples of Vr치ble neighbourhoods separately.

-   Vr치ble, Pareto + site map (exerpts)

-   One neighbourhood, same.

## Summary of findings

END Chapter
