# Methods: Fractal image analysis {#images-methods}

fdasdf

## Calculating fractal dimension and lacunarity

-   Box-counting, lit. @mancuso2021, @li2009, @klinkenberg1994

-   Gliding-box algorithms, @allain1991, @hingee2019, @cheng1997, @plotnick1996

-   Caveats:

    -   Fractional box-counting dimension does not equal self-similarity in a simple way

    -   My summary *L* is not equal to the one used in *FracLac* and thus by @farías-pelayo2017

-   Image preparation procedure

## Effects from image size and element count

Refer to Figure \@ref(fig:08-im-size) below.

(ref:08-im-size) Different image sizes

```{r 08-im-size, fig.cap="(ref:08-im-size)"}
load("Results/fig08_im_size.RData")
fig08_im_size
```

Relationship between image density (built-up area) and fractal dimension, evaluated in @thomas2007, where they show that these two parameters, under certain conditions (constant observation window, prefactor values close to 1), are exponentially correlated, which is also what I found in my article. They show that observation window size and shape, as well as centroid placement, have little influence on *D*, while they have more influence on density when the pattern is not homogeneous. They do show, however, that images with the same density may have quite much variation in *D*, which is reflected in the layouts. Judging from their examples, more clustered layouts give higher *D* values, while more dispersed or dusty layouts give lower *D*, when density is constant. I ignore the use of prefactor values. According to @thomas2007, density is a crude measure of the overall intensity of the pattern, while fractal dimension is characterises the morphological structure, though it is not directly descriptive.

![Selection of the different tested image sizes](Results/fig08_size_im.pdf){#fig:08-size-im}

Then on element count, Figure \@ref(fig:08-N).

(ref:08-N) Element count (N).

```{r 08-N, fig.cap="(ref:08-N)"}
load("Results/fig08_N.RData")
fig08_N
```

Then to the illustration on Figure \@ref(fig:08-N-im).

(ref:08-N-im)

![Caption](Results/fig08_N_im.pdf){#fig:08-N-im}

Hope this works.

## Effects from density and size distribution

rasdf. Refer to Figure \@ref(fig:08-dens).

(ref:08-dens) Different image densities

```{r 08-dens, fig.cap="(ref:08-dens)"}
load("Results/fig08_dens.RData")
fig08_dens
```

Then to illustrate, Figure \@ref(fig:08-dens-im).

![Different image densities](Results/fig08_dens_im.pdf){#fig:08-dens-im}

Then to size distributions. See Figure \@ref(fig:08-distr).

(ref:08-distr) Different size distributions of the elements contained in the images

```{r 08-distr, fig.cap="(ref:08-distr)"}
load("Results/fig08_distr.RData")
fig08_distr
```

Then refer to Figure \@ref(fig:08-distr-im).

![Different size distributions](Results/fig08_distr_im.pdf){#fig:08-distr-im}

And go on.

## Quantification of self-similarity and random noise

Now, what's really interesting. Self-similarity as clustering in Figure \@ref(fig:08-clustering).

(ref:08-clustering) Different degrees of spatial clustering and self-similarity

```{r 08-clustering, fig.cap="(ref:08-clustering)"}
load("Results/fig08_clustering.RData")
fig08_clustering
```

See the illustration on Figure \@ref(fig:08-clustering-im).

![Different degrees of clustering and self-similarity](Results/fig08_clustering_im.pdf){#fig:08-clustering-im}

And also random noise, see

(ref:08-noise) Different degrees of random spatial noise

```{r 08-noise, fig.cap="(ref:08-noise)"}
load("Results/fig08_noise.RData")
fig08_noise
```

Then see illustration on Figure \@ref(fig:08-noise).

![Different degrees of random spatial noise](Results/fig08_noise_im.pdf){#fig:08-noise-im}

That's it for the tests. Summarise below.

## Summary of procedure and tests

Just to sum up the different tests:

(ref:08-all) Results from all test series

```{r 08-all, fig.cap="(ref:08-all)"}
load("Results/fig08_all.RData")
fig08_all
```

Though the value ranges of the different tested factors are difficult to compare,... etc.

-   Procedure for archaeological samples, same as in article

-   Procedure for synthetic sample

END Chapter
