# Results: Image analysis {#images-results}

## Settlements

fasdfadfs

(ref:09-all) Maybe I'll move this one to the previous chapter?

```{r 09-all, fig.cap="(ref:09-all)"}
load("Results/fig09_all.RData")
fig09_all
```

fdsasdf

(ref:09-settle-points) fdsasfasd

```{r 09-settle-points, fig.cap="(ref:09-settle-points)"}
load("Results/fig09_settle_points.RData")
fig09_settle_points
```

fdsasdfasdf

![caption here](Results/fig09_settlements.pdf){#fig09-settlements}

jklølkjølkjgkh

## Quarters/neighbourhoods

fdasdfasdf

(ref:09-quart-points) fdsasdfad

```{r 09-quart-points, fig.cap="(ref:09-quart-points)"}
load("Results/fig09_quart_points.RData")
fig09_quart_points
```

fdasdfasd

![caption yuhuuu](Results/fig09_quarters.pdf){#fig09-quarters}

fdsasdfasdfghfghdfghf

## Temporal samples (Vráble)

fadsfasdf

(ref:09-time-points) fdsasdfasdfsdf

```{r 09-time-points, fig.cap="(ref:09-time-points)"}
load("Results/fig09_time_points.RData")
fig09_time_points
```

fdsasdsfdghs

![fdghfghrt](Results/fig09_time.pdf){#09-time}

fsfghertyrty

## Summary of findings

adfadfasdfa

END Chapter
