# (APPENDIX) Appendix {.unnumbered}

# Archaeological plan images {#append-archaeo}

The following plates include image series of archaeological settlement plans analysed in Chapter \@ref(images-results). The different image series are not to scale between them.

-   Settlements -- images are to scale within series

-   Quarters -- each plate is to scale internally, but not between each other

-   Time samples -- images are to scale within series

    -   Table \@ref(tab:append-time-tab) -- time samples, modelled sample dates and house orientations

See overview figures in Chapter \@ref(material) for absolute scales, and details on image preparation.

Original image files can be downloaded from project repository.

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/settlements.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/maidanetske.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/nebelivka.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/quart_neb.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/quart_vra.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/time_1.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/time_2.pdf")
```

\newpage

(ref:time-tab) Time samples, modelled sample dates, maximum and minimum house orientations of included houses. Details in Section \@ref(time-samples)

```{r append-time-tab}
load("Data/time_samples.RData") 
kable(time_samples, booktabs = TRUE, caption = "(ref:time-tab)", 
      caption.short = "Vráble time samples, modelled dates and house orientations") %>%  
  kable_styling(latex_options = c("striped", "scale down"))
```

# Synthetic images {#append-synth}

The following plates contain the synthetic image series used for analysis in Chapter \@ref(images-methods), in the following order (plates are read row-wise from top left):

-   Variable image size and resolution -- pattern is zoomed in as the image gets larger

-   Variable image size and observation window -- as image gets larger, a larger extent of the pattern becomes visible; resolution stays the same

-   Variable resolution and observation window -- pattern is zoomed out as larger extent is included; image size is fixed

-   Variable image density (from here, image size, resolution and observation window are constant)

-   Variable size distribution of elements

-   Variable level of hierarchical clustering

-   Variable level of added random noise

See chapter text for details.

Original image files and R code for generating them can be downloaded from project repository.

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/size_res.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/size_obs.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/res_obs.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/density.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/distr.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/clustering.pdf")
```

\newpage

```{r out.width = "100%"}
include_graphics("Appendix/noise.pdf")
```
