---
title: "Fractal Analysis as Method for studying Social Hierarchy in Prehistoric Settlement Plans"
subtitle: "with case studies from the Linear Pottery and Trypillia cultures (5.500 - 3.500 cal. BCE)"
author: 
  - "Hallvard Bruvoll"
  - "Department of Archaeology, Conservation and History"
  - "University of Oslo"
header-includes:
  - \usepackage{titling}
  - \pretitle{\begin{center}
    \includegraphics[width=2cm, height=2cm]{UiO_logo.png}\LARGE\\}
  - \posttitle{\end{center}}
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    toc_appendix: yes
#    citation_package: natbib
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
github-repo: hallvardbruvoll/Hallvard_Thesis_2
description: "This is from 'description' in the YAML"
geometry: "left=2.9cm, right=2.5cm, top=2.5cm, bottom=2.5cm"
mainfont: Times New Roman
fontsize: 12pt
linestretch: 1.5
toc: yes
toc-depth: 1
#lof: yes
#lot: yes
---

# Preface {-}

Thanks to everyone.

fdfasdf add citation @arponen2016.

This is a *sample* book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
